---
title: "Smoke Bioaerosol Concentration and Emission Mixed Models"
format:
  html:
    code-fold: true
    df-print: paged
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    smooth-scroll: true
    fig-cap-location: bottom
    embed-resources: true
    code-summary: "Show code"
    code-tools: true
  pdf: default
---
#| include: false

options(width = 110)
knitr::opts_chunk$set(
  dpi = 600,           # Higher DPI for better resolution
  fig.width = 7,      # Width in inches
  fig.height = 5,      # Height in inches
  out.width = "100%"   # How wide the image appears in the HTML
)

```{r}

library(tidyverse)
library(emmeans)
library(glmmTMB)
library(DHARMa)
```
# Konza 4 spores and bacteria--Tallgrass Prairie wildland fire microbial emissions
```{r}
spores_pa_C <- read_csv('./Output/Output_data/K4/k4_Spores_PA_C.csv') %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High")))

smoke_spores <- spores_pa_C %>%
  filter(SampleType == "Smoke")
  ```
  ## Spore concentration mixed Models
  ### Ambient vs. Smoke
  ### Smoke levels based on PM2.5 concentration ranges
  ```{r}
SporeCatModel <- glmmTMB(TotalSpores ~ SmokeLevel + offset(log(Slide_RepVolume_L)) + 
                           (1|SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = spores_pa_C, 
                         ziformula = ~1)
summary(SporeCatModel)

#| label: model-diagnostics
#| fig-cap: "Model diagnostic plots"

# Close all devices one at a time
while (length(dev.list()) > 0) {
  dev.off()
}

# Create new interactive window
windows()

# List all open plot devices
print(dev.list())

res <- simulateResiduals(fittedModel = SporeCatModel, plot = FALSE)

# Create diagnostic plots
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, spores_pa_C$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 


emm_log <- emmeans(SporeCatModel, ~ SmokeLevel)
emm_log

emm_response <- emmeans(SporeCatModel, ~ SmokeLevel, type = "response")
emm_response

rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")
rate_ratios_vs_none

spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
    )

# Create contrast dataframe automatically from emmeans output
contrast_df <- as.data.frame(rate_ratios_vs_none)

# Create significance bars 
sig_bars <- data.frame(
  x = c(1, 1),  # Both start from position 1 (None)
  xend = c(2, 3),  # End at positions 2 (Low) and 3 (High)
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

# Extract p-values using contrast names
sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

# Calculate maximum height for positioning significance bars
max_height <- max(spore_cat$CI_upper, na.rm = TRUE)

# Position bars at different heights to avoid overlap
sig_bars$y <- max_height * c(3.1, 3.3)

# Debugging: print contrast_df and sig_bars to check matching
print("contrast_df:")
print(contrast_df)
print("sig_bars:")
print(sig_bars)

# Ensure matching and correct label assignment
sig_bars$label <- sapply(sig_bars$pair, function(pair) {
  idx <- which(trimws(contrast_df$contrast) == trimws(pair))
  if (length(idx) > 0) {
    p <- contrast_df$p.value[idx]
    if (p < 0.001) {
      return("***")
    } else if (p < 0.01) {
      return("**")
    } else if (p < 0.05) {
      return("*")
    } else {
      return("ns")
    }
  } else {
    return("ns")
  }
})

# Debugging: print updated sig_bars
print("Updated sig_bars with labels:")
print(sig_bars)

# Add annotation for test and p-value in plotmath format for parse=TRUE
sig_bars$annotation <- paste0(
  "italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2)
)

#| label: spore-concentrations
#| fig-cap: "Spore concentrations by smoke level with significance bars"

# Create the main plot
p1 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = spores_pa_C, aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.1, color = "black", linewidth = 1.2) +
  labs(x = expression(paste("PM"[2.5], " μg/m"^3, " range")), 
       y = expression(paste("Spores/m"^3)),
       title = expression(paste("Wildland Fire vs. Ambient Tallgrass Prairie Fungal Spores by ", "PM"[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.7) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.02),
               linewidth = 0.7) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.02),
               linewidth = 0.7) +
  # Add annotation for test and p-value above the significance bar
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(≥400)")) +
  coord_cartesian(ylim = c(0, max(spores_pa_C$TotalSpores_FBLBcorr.m3 * 0.5)))

# Display the plot
print(p1)
```

```{r}
#| label: model-summary
#| echo: false

# Display model summary
summary(SporeCatModel)
```

```{r}
#| label: contrast-results
#| echo: false

# Display contrast results
print(rate_ratios_vs_none)
```
