---
title: "Smoke Bioaerosol Concentration and Emission Mixed Models"
format:
  html:
    code-fold: true
    df-print: paged
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    smooth-scroll: true
    fig-cap-location: bottom
    embed-resources: true
    code-summary: "Show code"
    code-tools: true
    margin-left: 2cm
    margin-right: 2cm
    code-block-width: 100%
  pdf:
    margin-left: 1.75cm
    margin-right: 1.75cm
    code-block-width: 100%
---

```{r setup, include=FALSE}
options(width = 100)  # Reduced from 110 to prevent text overflow
knitr::opts_chunk$set(
  dpi = 300,           # Higher DPI for better resolution
  fig.width = 6.5,      # Width in inches
  fig.height = 6,      # Height in inches
  out.width = "100%",  # How wide the image appears in the HTML
  warning = FALSE,      # Suppress all warnings in code chunks
  message = FALSE       # Suppress all messages in code chunks
)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(emmeans)
library(glmmTMB)
library(DHARMa)
library(patchwork)
library(ggpattern)
```

# [1] Study Overview

## [1.1] Site Description and Sampling Design

This study examined airborne microbes (fungi and bacteria) in ambient air and wildland fire smoke across two distinct ecosystems:

- **Utah subalpine** (FASMEE23): Subalpine forest fire regime
- **Kansas tallgrass prairie** (Konza4): Grassland fire regime

Samples were collected using drone-mounted equipment before fire ignition (ambient conditions) and during active burning (smoke conditions). Air and smoke samples were collected using filters and compensating pumps that drew specific volumes of air onto the sample filters.

## [1.2] Sample Collection Protocol

**Filter handling**: Before and after each sample collection, filters were placed and removed from sampling devices in situ using a mobile laboratory box to minimize air contamination and external influence on sample filters.

**Quality control**: Filter blanks were collected in situ to quantify contamination during filter handling and processing.

**Concurrent measurements**: Each drone flight simultaneously collected meteorological data, particulate matter (PM) concentrations, and carbon sensor data. *Note: Carbon sensor malfunctions resulted in incomplete carbon data for some samples.*

## [1.3] Microbial Enumeration Methods

Microbes were enumerated using epifluorescence microscopy after fluorescent staining:

**Counting protocol**:
- Magnification: 1000x
- Fields of view (FOV): 40 random locations per filter
- Filter area: ~226.98 mm²
- FOV area at 1000x: ~0.031 mm²
- Total possible FOVs per filter: 7,225

**Staining approaches**:
- **Bacteria**: Two-part viability stain allowing quantification of total bacteria, live bacteria, and dead bacteria
- **Fungi**: Single stain for spore enumeration (viability not distinguishable)

**Laboratory blanks**: Lab blanks were processed with each batch of staining solutions to account for contamination introduced during the staining process.

# [2] Research Questions

1. **Microbial abundance ambient vs. smoke intensity**: How do microbial concentrations in ambient air differ from concentrations in smoke of varying intensity?
   - Smoke intensity classified using PM<sub>2.5</sub> concentrations from onboard sensors

2. **Environmental drivers**: Do smoke-associated microbes correlate with meteorological variables or carbon concentrations?

3. **Bacterial viability**: Does the ratio of live to total bacteria differ between ambient and smoke conditions?

4. **Site comparison**: Are there differences in microbial abundances between study sites under both ambient and smoke conditions?

5. **Emission factors**: Using the carbon data, we can transform our microbe counts into the number of microbes emitted by a certain mass of fuel burning (e.g. #cells emitted per kilogram of biomass burning)---Known as emission factors (EFs)---Do EFs correlate with combustion efficiency (smoldering vs. flaming combustion)?

# [3] Statistical Analysis Approach

## [3.1] Generalized Linear Mixed Models (GLMMs)

**Response variable**: FOV count data (Research Questions 1-4)
**Offset**: log(volume sampled) to account for varying sample volumes

**Random effects structure**:
- Random intercept for SampleID (accounts for multiple FOVs from the same sample)
- Random intercept for Lab Blank batch (accounts for batch-specific contamination levels during sample processing)

*Note: Research Question 5 (emission factors) requires a modified modeling approach incorporating:
- Averaging FOV counts across the 40 FOVs, scaling to the total filter area, and then normalizing to the air volume sampled and transforming based on carbon data.*

# [4] Konza4 spore data summary

```{r data-import, include=FALSE}

# Konza4 Spores
K4_spores_raw <- read_csv('./Output/Output_data/K4/k4_Spores_PA_C.csv')
Konza4_spores_pa_C <- K4_spores_raw %>%
  rename(Sample = Sample_num, TotalSpores_corrected.m3 = TotalSpores_FBLBcorr.m3, RepVolume_L = Slide_RepVolume_L) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High")),
    MeanVolume_m3 = mean(RepVolume_m3)
  ) %>%
  filter(SampleID != "14A")

K4_UAS <- read_csv('./Output/Output_data/K4/K4_FlightsAGL.csv') %>%  
  mutate(
    datetime.utc. = ymd_hms(datetime.utc., tz = ('UTC'))
  ) %>%
  group_by(Sample) %>%
  arrange(datetime.utc., .by_group = TRUE) %>%
  mutate(
    min_time = min(datetime.utc.),
    max_time = max(datetime.utc.)
  ) %>%
  filter( # filtering flights to times two minutes after take-off and two minutes before landing
    datetime.utc. >= min_time + minutes(2),
    datetime.utc. <= max_time - minutes(2)
  )

K4_UAS_samples <- K4_UAS %>%
  group_by(Sample) %>%
  summarise(
    MinAGL = min(AGL_m),
    MeanAGL = mean(AGL_m),
    MedianAGL = median(AGL_m),
    MaxAGL = max(AGL_m)
    )

Konza4_spores_pa_C <- left_join(Konza4_spores_pa_C, K4_UAS_samples, by = "Sample")

# Konza4 Bacteria
K4_bacteria_raw <- read_csv('./Output/Output_data/K4/k4_Bacteria_PA_C.csv') 

K4_bacteria <- K4_bacteria_raw %>%
  select(-Sample) %>%
  rename(Sample = Sample_num, TotalCells_corrected.m3 = TotalCells_FBLBcorr.m3) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  mutate(
    Sample = as.numeric(Sample),
    MeanVolume_m3 = mean(RepVolume_m3)
  )

K4_bacteria <- left_join(K4_bacteria, K4_UAS_samples, by = "Sample")

# FASMEE23 Spores
FASMEE23_spores_raw <- read.csv('./Output/Output_data/FASMEE23/FASMEE23_Spores_PA_C.csv') 
FASMEE23_spores_pa_C <- FASMEE23_spores_raw %>%
  rename(TotalSpores_corrected.m3 = TotalSpores_FBLBcorr.m3, ALL_carbon_mg.m3 = Gas_TC_Carbon_mg.m3) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High")),
    MeanVolume_m3 = mean(RepVolume_m3)
  )

F23_BlueUAS <- read_csv('./Output/Output_data/FASMEE23/FASMEE23_FlightsAGL.csv') %>%  
  mutate(
    datetime.utc. = ymd_hms(datetime.utc., tz = ('UTC'))
  ) %>%
  group_by(Sample) %>%
  arrange(datetime.utc., .by_group = TRUE) %>%
  mutate(
    min_time = min(datetime.utc.),
    max_time = max(datetime.utc.)
  ) %>%
  filter( # filtering flights to times two minutes after take-off and two minutes before landing
    datetime.utc. >= min_time + minutes(2),
    datetime.utc. <= max_time - minutes(2)
  )

F23_BlueUAS_samples <- F23_BlueUAS %>%
  group_by(SampleID_short) %>%
  summarise(
    MinAGL = min(AGL_m),
    MeanAGL = mean(AGL_m),
    MedianAGL = median(AGL_m),
    MaxAGL = max(AGL_m)
    )

F23_BlueSpores <- FASMEE23_spores_pa_C %>%
  filter(Platform == "Blue") %>%
  left_join(., F23_BlueUAS_samples, by = "SampleID_short")

# FASMEE23 Bacteria
FASMEE23_bacteria_raw <- read_csv('./Output/Output_data/FASMEE23/FASMEE23_Bacteria_PA_C.csv') 
F23_bacteria <- FASMEE23_bacteria_raw %>%
  rename(TotalCells_corrected.m3 = TotalCells_FBLBcorr.m3, ALL_carbon_mg.m3 = Gas_TC_Carbon_mg.m3) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High")),
    LiveRatio = LIVECounts / TotalCells,
    MeanVolume_m3 = mean(RepVolume_m3)    
  )

F23_BlueBacteria  <- F23_bacteria %>%
  filter(Platform == "Blue") %>%
  left_join(., F23_BlueUAS_samples, by = "SampleID_short")

# FASMEE23 qPCR
F23_qPCR <- read_csv('./Output/Output_data/FASMEE23/FASMEE23_qPCR_SampleInfo.csv') %>%
  select(-Platform, -SampleID)

# Smoke only samples
Konza4_smoke_spores <- Konza4_spores_pa_C %>%
  filter(SampleType == "Smoke")

FASMEE23_smoke_spores <- FASMEE23_spores_pa_C %>%
  filter(SampleType == "Smoke" & Platform == "Blue")

FASMEE23_smoke_bacteria <- F23_bacteria %>%
  filter(SampleType == "Smoke" & Platform == "Blue")

K4_FBLB_spores <- read_csv('./Output/Output_data/K4/k4_FBLB_spores.csv') %>% 
    mutate(
    LB_Batch = as.factor(LB_Batch),
    SampleType = as.factor(SampleType)) 

K4_FBLB_spore_stat <- K4_FBLB_spores %>%
  group_by(SampleType) %>%
  summarise(
    medianSpores.fov = median(TotalSpores)
  ) %>% ungroup %>%
  pivot_wider(names_from = SampleType, values_from = medianSpores.fov)

K4_FBLB_bacteria <- read_csv('./Output/Output_data/K4/k4_FBLB_bacteria.csv') %>% 
    mutate(
    LB_Batch = as.factor(LB_Batch),
    SampleType = as.factor(SampleType))

K4_FB_bacteria_stat <- K4_FBLB_bacteria %>%
  filter(SampleType == "FieldBlank") %>%
  group_by(SampleType) %>%
  summarise(
    median_TotBacteria.fov = median(TotalCells),
    median_LiveBacteria.fov = median(LIVECounts),
    median_DeadBacteria.fov = median(DEADCounts)
  )

K4_LB_bacteria_stat <- K4_FBLB_bacteria %>%
  filter(SampleType == "LabBlank") %>%
  group_by(SampleType) %>%
  summarise(
    median_TotBacteria.fov = median(TotalCells),
    median_LiveBacteria.fov = median(LIVECounts),
    median_DeadBacteria.fov = median(DEADCounts)
  )

F23_FBLB_bacteria <- read_csv('./Output/Output_data/FASMEE23/F23_FBLB_bacteria.csv') %>% 
    mutate(
    LB_Batch = as.factor(LB_Batch),
    SampleType = as.factor(SampleType)) 

F23_FB_bacteria_stat <- F23_FBLB_bacteria %>%
  filter(SampleType == "FieldBlank") %>%
  group_by(SampleType) %>%
  summarise(
    Median_TotBacteria.fov = median(TotalCells),
    Median_LiveBacteria.fov = median(LIVECounts),
    Median_DeadBacteria.fov = median(DEADCounts)
  )

F23_LB_bacteria_stat <- F23_FBLB_bacteria %>%
  filter(SampleType == "LabBlank") %>%
  group_by(SampleType) %>%
  summarise(
    median_TotBacteria.fov = median(TotalCells),
    median_LiveBacteria.fov = median(LIVECounts),
    median_DeadBacteria.fov = median(DEADCounts)
  )

F23_FBLB_spores <- read_csv('./Output/Output_data/FASMEE23/F23_FBLB_spores.csv') %>% 
    mutate(
    LB_Batch = as.factor(LB_Batch),
    SampleType = as.factor(SampleType))

F23_FBLB_spore_stat <- F23_FBLB_spores %>%
  group_by(SampleType) %>%
  summarise(
    medianSpores.fov = median(TotalSpores)
  ) %>% ungroup %>%
  pivot_wider(names_from = SampleType, values_from = medianSpores.fov)
```

```{r data-load, echo=FALSE}

# Tidy data summary
cat("=== Konza4 Spore Data Summary ===\n")
cat("Dataset dimensions:", nrow(Konza4_spores_pa_C), "rows ×", ncol(Konza4_spores_pa_C), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(Konza4_spores_pa_C$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "RepVolume_L", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(Konza4_spores_pa_C[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(Konza4_spores_pa_C[key_vars], function(x) round(mean(x, na.rm = TRUE), 2)),
  SD = sapply(Konza4_spores_pa_C[key_vars], function(x) round(sd(x, na.rm = TRUE), 2)),
  Min = sapply(Konza4_spores_pa_C[key_vars], function(x) round(min(x, na.rm = TRUE), 2)),
  Q25 = sapply(Konza4_spores_pa_C[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 2)),
  Median = sapply(Konza4_spores_pa_C[key_vars], function(x) round(median(x, na.rm = TRUE), 2)),
  Q75 = sapply(Konza4_spores_pa_C[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 2)),
  Max = sapply(Konza4_spores_pa_C[key_vars], function(x) round(max(x, na.rm = TRUE), 2))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(Konza4_spores_pa_C$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(Konza4_spores_pa_C$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(Konza4_spores_pa_C$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(Konza4_spores_pa_C$LB_Batch), "\n")
cat("- Sample types:", paste(unique(Konza4_spores_pa_C$SampleType), collapse = ", "), "\n")

cat("\n=== Data Distribution Plots ===\n")

create_hist_density <- function(data, var_name, title, x_label) {
  plot_data <- data[!is.na(data[[var_name]]), ]
  
  ggplot(plot_data, aes(x = .data[[var_name]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 15, fill = "#0072B2", alpha = 0.7, color = "black") +
    geom_density(color = "#D55E00", linewidth = 1.2) +
    labs(title = title, x = x_label, y = "Density") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
          axis.text = element_text(size = 9),
          axis.title = element_text(size = 9))
}

# Create plots for each variable
p1 <- create_hist_density(Konza4_spores_pa_C, "TotalSpores", "Total Spores Distribution", "Total Spores")
p2 <- create_hist_density(Konza4_spores_pa_C, "TotalSpores_corrected.m3", "Total Spore concentration Distribution", "Total spore concentration (cells/m³)")
p3 <- create_hist_density(Konza4_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p4 <- create_hist_density(Konza4_spores_pa_C, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p5 <- create_hist_density(Konza4_spores_pa_C, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p6 <- create_hist_density(Konza4_spores_pa_C, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p7 <- create_hist_density(Konza4_spores_pa_C, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p8 <- create_hist_density(Konza4_spores_pa_C, "MeanMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) + 
  plot_annotation(
    title = "Konza4 Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )

# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")
boxplot_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")

create_boxplot <- function(data, var_name, title) {
  ggplot(data, aes(x = SmokeLevel, y = .data[[var_name]], fill = SmokeLevel)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    labs(title = title, x = "Smoke Level", y = var_name) +
    scale_fill_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
          legend.position = "none",
          axis.text = element_text(size = 9),
          axis.title = element_text(size = 9))
}

bp1 <- create_boxplot(Konza4_spores_pa_C, "TotalSpores_corrected.m3", "Spore Concentration by Smoke Level")
bp2 <- create_boxplot(Konza4_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp3 <- create_boxplot(Konza4_spores_pa_C, "MeanTemp_C", "Temperature by Smoke Level")
bp4 <- create_boxplot(Konza4_spores_pa_C, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")
library(corrplot)

# Select numeric variables for correlation
cor_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")
cor_data <- Konza4_spores_pa_C[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Spores vs PM2.5
sp1 <- ggplot(Konza4_spores_pa_C, aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spores vs Carbon
sp2 <- ggplot(Konza4_spores_pa_C, aes(x = ALL_carbon_mg.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs RH
sp3 <- ggplot(Konza4_spores_pa_C, aes(x = MeanRH, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs Relative Humidity", 
       x = "Relative Humidity (%)", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs MCE
sp4 <- ggplot(Konza4_spores_pa_C, aes(x = MeanMCE, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs MCE", 
       x = "Modified Combustion Efficiency", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1) / (sp2) / (sp3) / (sp4) +
  plot_annotation(title = "Key Variable Relationships")


# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- Konza4_spores_pa_C %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

```{r}
#| eval: false
#| include: false
 
  
# K4_SporeLabBlanks_m <- glmmTMB(
#   TotalSpores ~ LB_Batch + (1|SampleID),
#   family = nbinom2(link = "log"), 
#   data = K4_SporeLabBlanks, 
#   ziformula = ~-1)
# summary(K4_SporeLabBlanks_m )
# 
# res <- simulateResiduals(fittedModel = K4_SporeLabBlanks_m , plot = FALSE)
# plotQQunif(res)
# testCategorical(res, catPred = K4_SporeLabBlanks$LB_Batch) # Residuals by group
# plotResiduals(res)
# testZeroInflation(res)
# simOutput_cond <- simulateResiduals(K4_SporeLabBlanks_m , re.form = NULL)
# testDispersion(simOutput_cond) 
# 
# print(drop1(K4_SporeLabBlanks_m , test = "Chisq"))
# 
# emm_log_K4LBspore <- emmeans(K4_SporeLabBlanks_m , ~ LB_Batch)
# emm_response_K4LBspore <- emmeans(K4_SporeLabBlanks_m, ~ LB_Batch, type = "response")
# rate_ratios_K4LBspore <- pairs(emm_response_K4LBspore, reverse = T)
# 
# 
# K4_BacteriaLabBlanks_m <- glmmTMB(
#   TotalCells ~ LB_Batch + SampleType + (1|SampleID),
#   family = nbinom2(link = "log"), 
#   dispformula = ~SampleType + (1|SampleID),
#   data = K4_BacteriaLabBlanks)
# summary(K4_BacteriaLabBlanks_m)
# 
# res <- simulateResiduals(fittedModel = K4_BacteriaLabBlanks_m , plot = FALSE)
# plotQQunif(res)
# testCategorical(res, catPred = K4_BacteriaLabBlanks$LB_Batch) # Residuals by group
# plotResiduals(res)
# testZeroInflation(res)
# simOutput_cond <- simulateResiduals(K4_BacteriaLabBlanks_m , re.form = NULL)
# testDispersion(simOutput_cond) 
# 
# print(drop1(K4_BacteriaLabBlanks_m , test = "Chisq"))
# 
# 
# emm_log_K4LBbacteria <- emmeans(K4_BacteriaLabBlanks_m , ~ LB_Batch)
# emm_response_K4LBbacteria <- emmeans(K4_BacteriaLabBlanks_m, ~ SampleType, type = "response")
# rate_ratios_K4LBbacteria <- pairs(emm_response_K4LBbacteria, reverse = T)
# 
# # FASMEE23 bacteria lab blanks
# 
# F23_BacteriaLabBlanks_m <- glmmTMB(
#   TotalCells ~ SampleType + (1|SampleID) + (1|LB_Batch),
#   family = nbinom2(link = "log"), 
#   ziformula = ~SampleType,
#   data = F23_BacteriaLabBlanks)
# summary(F23_BacteriaLabBlanks_m)
# 
# res <- simulateResiduals(fittedModel = F23_BacteriaLabBlanks_m , plot = FALSE)
# plotQQunif(res)
# testCategorical(res, catPred = F23_BacteriaLabBlanks$SampleType) # Residuals by group
# plotResiduals(res)
# testZeroInflation(res)
# simOutput_cond <- simulateResiduals(F23_BacteriaLabBlanks_m , re.form = NULL)
# testDispersion(simOutput_cond) 
# 
# print(drop1(F23_BacteriaLabBlanks_m , test = "Chisq"))
# 
# 
# emm_log_F23LBbacteria <- emmeans(F23_BacteriaLabBlanks_m , ~ LB_Batch)
# emm_response_F23LBbacteria <- emmeans(F23_BacteriaLabBlanks_m, ~ SampleType, type = "response")
# rate_ratios_F23LBbacteria <- pairs(emm_response_F23LBbacteria, reverse = T)


```


# [5] Modeling for Research Question #1: Konza4 spore abundance in ambient vs. smoke conditions

## [5.1] GLMM model for Konza4 spores and DHARMa diagnostics for assesment 
```{r K4 spores vs smoke levels}
Konza4_SporeCatModel <- glmmTMB(
  TotalSpores ~ SmokeLevel + offset(log(RepVolume_L)) + 
    (1|SampleID) + (1|LB_Batch),
  family = nbinom2(link = "log"), 
  data = Konza4_spores_pa_C, 
  ziformula = ~-1)
summary(Konza4_SporeCatModel)

res <- simulateResiduals(fittedModel = Konza4_SporeCatModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = Konza4_spores_pa_C$SmokeLevel) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(Konza4_SporeCatModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(Konza4_SporeCatModel, test = "Chisq"))

#Calculate marginal means and contrasts
emm_log <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel)
emm_response <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
spore_cat <- as.data.frame(emm_response) %>%
  mutate(
    Spores.m3 = ((response-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032,
    CI_lower = ((asymp.LCL-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032,
    CI_upper = ((asymp.UCL-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032
  )

print(spore_cat)

```
#### Smoke level is not a significant predictor (p = 0.09899), so lets see if smoke vs. ambient is 

```{r}

Konza4_Spore_SampleTypeModel <- glmmTMB(
  TotalSpores ~ SampleType + offset(log(RepVolume_L)) + 
    (1|SampleID) + (1|LB_Batch),
  family = nbinom2(link = "log"), 
  data = Konza4_spores_pa_C, 
  ziformula = ~-1)
summary(Konza4_Spore_SampleTypeModel)

res <- simulateResiduals(fittedModel = Konza4_Spore_SampleTypeModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = Konza4_spores_pa_C$SampleType) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(Konza4_Spore_SampleTypeModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(Konza4_Spore_SampleTypeModel, test = "Chisq"))

# Calculate marginal means and contrasts
emm_log <- emmeans(Konza4_Spore_SampleTypeModel, ~ SampleType)
emm_response <- emmeans(Konza4_Spore_SampleTypeModel, ~ SampleType, type = "response")
K4_rate_ratios <- pairs(emm_response, reverse = T)

# Create prediction dataframe
K4_spore_sampletype <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = ((response-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032,
    CI_lower = ((asymp.LCL-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032,
    CI_upper = ((asymp.UCL-(K4_FBLB_spore_stat$LabBlank + K4_FBLB_spore_stat$FieldBlank))*7225)/0.032
  )

```

## [5.2] Predictions from Konza4 spore GLMM model
```{r Marginal means, echo=FALSE}
print(K4_spore_sampletype)
```

## [5.3] Ratios from Konza4 spore GLMM model
```{r contrast-results, echo=FALSE}
print(summary(K4_rate_ratios, infer = T))
```

## [5.4] Plotting Konza4 spore model results with raw data as points/violin plots
```{r Plotting model results, fig.cap="Konza 4 Spore GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
K4_contrast_df <- as.data.frame(K4_rate_ratios)
K4_sig_bars <- data.frame(
  x = c(1),
  xend = c(2),
  pair = c("Smoke / Ambient"),
  stringsAsFactors = FALSE
)

K4_sig_bars$p_value <- sapply(K4_sig_bars$pair, function(pair) {
  idx <- which(K4_contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(K4_contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

K4_max_height <- max(K4_spore_sampletype$CI_upper, na.rm = TRUE)
K4_sig_bars$y <- K4_max_height * c(2.7)
K4_sig_bars$annotation <- ifelse(
  is.na(K4_sig_bars$p_value),
  'plain("")',
  paste0("italic(p) ==", signif(K4_sig_bars$p_value, 2))
)

k4_spore_plt <- ggplot(K4_spore_sampletype, aes(x = SampleType, y = Spores.m3)) + 
  geom_point(data = Konza4_spores_pa_C, 
             aes(x = SampleType, y = TotalSpores_corrected.m3, color = "Raw Data"), 
             position = position_jitter(width = 0.15), 
             alpha = 0.5, size = 1) +
  geom_violin(data = Konza4_spores_pa_C, 
              aes(x = SampleType, y = TotalSpores_corrected.m3), 
              fill = NA, color = "gold3", width = 0.8) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "blue2", alpha = 0.6, linewidth = 0.8) +
  geom_point(aes(y = Spores.m3, color = "Model Output"), size = 1.5) +
  geom_segment(data = K4_sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = K4_sig_bars,
               aes(x = x, xend = x, y = y, yend = y - K4_max_height * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = K4_sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - K4_max_height * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_text(data = K4_sig_bars,
            aes(x = (x + xend) / 2, y = y + K4_max_height * 0.02, label = annotation),
            size = 2, vjust = 0, parse = TRUE) +
  geom_segment(aes(x = as.numeric(SampleType), 
                   xend = as.numeric(SampleType) + 0.2,  
                   y = Spores.m3, 
                   yend = Spores.m3 + 60000),  
               linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_label(data = K4_spore_sampletype, 
            aes(label = paste0("hat(mu) == ", 
                              round(Spores.m3/10^floor(log10(Spores.m3)), 2), 
                              " %*% 10^", floor(log10(Spores.m3))),
                x = as.numeric(SampleType) + 0.2,  
                y = Spores.m3 + 60000),
            hjust = 0, size = 2, parse = TRUE,
            fill = "white", color = "black", alpha = 0.3, 
            label.padding = unit(0.25, "lines"),
            label.r = unit(0.15, "lines")) +
  scale_color_manual(values = c("Model Output" = "blue2", "Raw Data" = "gold3")) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 4, alpha = 1))) +
  scale_y_continuous(labels = scales::scientific, 
                   limits = c(0, max(Konza4_spores_pa_C$TotalSpores_corrected.m3 * 0.5))) +
  labs(x = "Sample Type", 
       y = expression("Tallgrass prairie: spores"~m^{-3}),
       color = "") +
  theme_bw() +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        panel.grid = element_blank(),
        legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.key = element_rect(fill = "white", color = NA)) 

print(k4_spore_plt)

ggsave('Output/Output_figs/K4/K4_SmokeVsAmbient_SporeModel.png', k4_spore_plt, dpi = 600, width = 5, height = 4, units = "in")


```

# [6] FASMEE23 spore data summary
```{r FASMEE23 spores, echo=FALSE}
# Data already loaded and wrangled in data-import chunk

# Tidy data summary
cat("=== FASMEE23 Spore Data Summary ===\n")
cat("Dataset dimensions:", nrow(FASMEE23_spores_pa_C), "rows ×", ncol(FASMEE23_spores_pa_C), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(FASMEE23_spores_pa_C$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "RepVolume_L", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(FASMEE23_spores_pa_C[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(mean(x, na.rm = TRUE), 1)),
  SD = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(sd(x, na.rm = TRUE), 1)),
  Min = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(min(x, na.rm = TRUE), 1)),
  Q25 = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 1)),
  Median = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(median(x, na.rm = TRUE), 1)),
  Q75 = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 1)),
  Max = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(max(x, na.rm = TRUE), 1))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(FASMEE23_spores_pa_C$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(FASMEE23_spores_pa_C$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(FASMEE23_spores_pa_C$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(FASMEE23_spores_pa_C$LB_Batch), "\n")
cat("- Sample types:", paste(unique(FASMEE23_spores_pa_C$SampleType), collapse = ", "), "\n")
cat("- Platforms:", paste(unique(FASMEE23_spores_pa_C$Platform), collapse = ", "), "\n")

# Data Distribution Plots
cat("\n=== Data Distribution Plots ===\n")

# Create plots for each variable
p1 <- create_hist_density(FASMEE23_spores_pa_C, "TotalSpores", "Total Spores Distribution", "Total Spores")
p2 <- create_hist_density(FASMEE23_spores_pa_C, "TotalSpores_corrected.m3", "Total Spore concentration Distribution", "Total spore concentration (cells/m³)")
p3 <- create_hist_density(FASMEE23_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p4 <- create_hist_density(FASMEE23_spores_pa_C, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p5 <- create_hist_density(FASMEE23_spores_pa_C, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p6 <- create_hist_density(FASMEE23_spores_pa_C, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p7 <- create_hist_density(FASMEE23_spores_pa_C, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p8 <- create_hist_density(FASMEE23_spores_pa_C, "MeanMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) + 
  plot_annotation(
    title = "FASMEE23 Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )

# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")

bp1 <- create_boxplot(FASMEE23_spores_pa_C, "TotalSpores_corrected.m3", "Spore Concentration by Smoke Level")
bp2 <- create_boxplot(FASMEE23_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp3 <- create_boxplot(FASMEE23_spores_pa_C, "MeanTemp_C", "Temperature by Smoke Level")
bp4 <- create_boxplot(FASMEE23_spores_pa_C, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")

# Select numeric variables for correlation
cor_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")
cor_data <- FASMEE23_spores_pa_C[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Spores vs PM2.5
sp1 <- ggplot(FASMEE23_spores_pa_C, aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spores vs Carbon
sp2 <- ggplot(FASMEE23_spores_pa_C, aes(x = ALL_carbon_mg.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs RH
sp3 <- ggplot(FASMEE23_spores_pa_C, aes(x = MeanRH, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs Relative Humidity", 
       x = "Relative Humidity (%)", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs MCE
sp4 <- ggplot(FASMEE23_spores_pa_C, aes(x = MeanMCE, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs MCE", 
       x = "Modified Combustion Efficiency", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1) / (sp2) / (sp3) / (sp4) +
  plot_annotation(title = "Key Variable Relationships")

# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- FASMEE23_spores_pa_C %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

# [7] Modeling for Research Question #1: FASMEE23 spore abundance in ambient vs. smoke conditions

## [7.1] GLMM model for FASMEE23 spores and DHARMa diagnostics for assessment

```{r FASMEE23 spore ambient vs smoke}
FASMEE23_SporeCatModel <- glmmTMB(TotalSpores ~ SampleType + offset(log(RepVolume_L)) + 
                          (1|Platform:SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_spores_pa_C, 
                        ziformula = ~-1, dispformula = ~ (1|Platform:SampleID))
summary(FASMEE23_SporeCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_SporeCatModel, plot = FALSE)
plotResiduals(res, FASMEE23_spores_pa_C$RepVolume_L)
plotQQunif(res)
testCategorical(res, catPred = FASMEE23_spores_pa_C$SampleType) # Residuals by group

testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_SporeCatModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(FASMEE23_SporeCatModel, test = "Chisq"))

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_SporeCatModel, ~ SampleType)
emm_response <- emmeans(FASMEE23_SporeCatModel, ~ SampleType, type = "response")
F23_rate_ratios <- pairs(emm_response, reverse = T)

# Create prediction dataframe
F23_spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = ((response - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025,
    CI_lower = ((asymp.LCL - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025,
    CI_upper = ((asymp.UCL - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025
  )

print(F23_spore_cat)

print(summary(F23_rate_ratios, infer = T))

```

```{r, FASMEE23 Spores Smoke Vs Ambient Plot}

# Create significance bars
F23_spores_contrast_df <- as.data.frame(F23_rate_ratios)
F23_spores_sig_bars <- data.frame(
  x = c(1),
  xend = c(2),
  pair = c("Smoke / Ambient"),
  stringsAsFactors = FALSE
)

F23_spores_sig_bars$p_value <- sapply(F23_spores_sig_bars$pair, function(pair) {
  idx <- which(F23_spores_contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(F23_spores_contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

F23_spores_max_height <- max(F23_spore_cat$CI_upper, na.rm = TRUE)
F23_spores_sig_bars$y <- F23_spores_max_height * c(3.5)
F23_spores_sig_bars$annotation <- ifelse(
  is.na(F23_spores_sig_bars$p_value),
  'plain("")',
  paste0("italic(p) ==", signif(F23_spores_sig_bars$p_value, 2))
)

F23_spore_plt <- ggplot(F23_spore_cat, aes(x = SampleType, y = Spores.m3)) + 
  geom_point(data = FASMEE23_spores_pa_C, 
             aes(x = SampleType, y = TotalSpores_corrected.m3, color = "Raw Data"), 
             position = position_jitter(width = 0.15), 
             alpha = 0.5, size = 1) +
  geom_violin(data = FASMEE23_spores_pa_C, 
              aes(x = SampleType, y = TotalSpores_corrected.m3), 
              fill = NA, color = "gold3", width = 0.8) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "blue2", alpha = 0.6, linewidth = 0.8) +
  geom_point(aes(y = Spores.m3, color = "Model Output"), size = 1.5) +
  geom_segment(data = F23_spores_sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = F23_spores_sig_bars,
               aes(x = x, xend = x, y = y, yend = y - F23_spores_max_height * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = F23_spores_sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - F23_spores_max_height * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_text(data = F23_spores_sig_bars,
            aes(x = (x + xend) / 2, y = y + F23_spores_max_height * 0.02, label = annotation),
            size = 2, vjust = 0, parse = TRUE) +
  geom_segment(aes(x = as.numeric(SampleType), 
                   xend = as.numeric(SampleType) + 0.18,  
                   y = Spores.m3, 
                   yend = Spores.m3 + 120000),  
               linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_label(data = F23_spore_cat, 
            aes(label = paste0("hat(mu) == ", 
                              round(Spores.m3/10^floor(log10(Spores.m3)), 2), 
                              " %*% 10^", floor(log10(Spores.m3))),
                x = as.numeric(SampleType) + 0.18,  
                y = Spores.m3 + 120000),
            hjust = 0, size = 2, parse = TRUE,
            fill = "white", color = "black", alpha = 0.3, 
            label.padding = unit(0.25, "lines"),
            label.r = unit(0.15, "lines")) +
  scale_color_manual(values = c("Model Output" = "blue2", "Raw Data" = "gold3")) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 4, alpha = 1))) +
  scale_y_continuous(labels = scales::scientific, 
                     limits = c(0, max(FASMEE23_spores_pa_C$TotalSpores_corrected.m3 * 0.45))) +
  labs(x = "Sample Type", 
       y = expression("Subalpine forest: spores"~m^{-3}),
       color = "") +
  theme_bw() +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        panel.grid = element_blank(),
        legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.key = element_rect(fill = "white", color = NA)) 

print(F23_spore_plt)

ggsave('Output/Output_figs/FASMEE23/F23_SmokeVsAmbient_SporeModel.png', F23_spore_plt, dpi = 600, width = 5, height = 4, units = "in")


```

```{r FASMEE23 spore smoke level GLMM}
F23_Spore_smokelvl_model <- glmmTMB(TotalSpores ~ SmokeLevel + offset(log(RepVolume_L)) + 
                          (1|Platform:SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_spores_pa_C, 
                        ziformula = ~-1, dispformula = ~ SmokeLevel)
summary(F23_Spore_smokelvl_model)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = F23_Spore_smokelvl_model, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = FASMEE23_spores_pa_C$SmokeLevel) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(F23_Spore_smokelvl_model, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(F23_Spore_smokelvl_model, test = "Chisq"))

# Calculate marginal means and contrasts
emm_log <- emmeans(F23_Spore_smokelvl_model, ~ SmokeLevel)
emm_response <- emmeans(F23_Spore_smokelvl_model, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- pairs(emm_response, reverse = T)

# Create prediction dataframe
spore_smokelvl <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = ((response - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025,
    CI_lower = ((asymp.LCL - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025,
    CI_upper = ((asymp.UCL - (F23_FBLB_spore_stat$FieldBlank + F23_FBLB_spore_stat$FieldBlank)) * 7225) / 0.025
  )

```

## [7.2] Predictions from FASMEE23 spore GLMM model
```{r FASMEE23 Marginal means, echo=FALSE}
print(spore_smokelvl)
```

## [7.3] Ratios from FASMEE23 spore GLMM model
```{r FASMEE23 contrast-results, echo=FALSE}
print(summary(rate_ratios_vs_none, infer = T))
```

## [7.4] Plotting FASMEE23 spore model results with raw data as points/violin plots
```{r FASMEE23 Plotting model results, fig.cap="FASMEE23 Spore GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1, 2),
  xend = c(2, 3, 3),
  pair = c("Low / None", "High / None", "High / Low"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(spore_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(2.75, 3.05, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(p)[Tukey-adj]==", signif(sig_bars$p_value, 2))
)

p2 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = FASMEE23_spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = FASMEE23_spores_pa_C, 
              aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~"("*mu*g~m^{-3}*") range"),  
       y = expression(Spores~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Subalpine Forest Fungal Spores by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2.07, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  # Add red points for means
  geom_point(aes(y = Spores.m3), color = "darkred", size = 3) +
  # Add dashed line from point to label
  geom_segment(aes(x = as.numeric(SmokeLevel), 
                   xend = as.numeric(SmokeLevel) + 0.15,  
                   y = Spores.m3, 
                   yend = Spores.m3 + 200000),  
               linetype = "dotdash", color = "darkred", linewidth = 0.7) +
  # Add lambda-hat labels
  geom_label(data = spore_cat, 
            aes(label = paste0("hat(mu) == ", 
                              round(Spores.m3/10^floor(log10(Spores.m3)), 2), 
                              " %*% 10^", floor(log10(Spores.m3))),
                x = as.numeric(SmokeLevel) + 0.15,  
                y = Spores.m3 + 200000),
            hjust = 0, size = 2.7, parse = TRUE,
            fill = "white", color = "black", 
            label.padding = unit(0.25, "lines"),
            label.r = unit(0.15, "lines")) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(FASMEE23_spores_pa_C$TotalSpores_corrected.m3 * 0.5)))

print(p2)

ggsave('Output/Output_figs/FASMEE23/FASMEE23_SmokeLevels_SporeModel.png', p2, dpi = 600, width = 6.5, height = 4, units = "in")
```

# [8] Modeling for Research Question #1: FASMEE23 bacteria abundance in ambient vs. smoke conditions

```{r FASMEE23 bacteria, echo=FALSE, fig.width=8, fig.height=10}
# Data already loaded and wrangled in data-import chunk

# Tidy data summary
cat("=== FASMEE23 Bacteria Data Summary ===\n")
cat("Dataset dimensions:", nrow(F23_bacteria), "rows ×", ncol(F23_bacteria), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(F23_bacteria$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalCells", "LIVECounts", "TotalCells_corrected.m3", "LiveRatio", "RepVolume_L", 
              "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(F23_bacteria[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(F23_bacteria[key_vars], function(x) round(mean(x, na.rm = TRUE), 1)),
  SD = sapply(F23_bacteria[key_vars], function(x) round(sd(x, na.rm = TRUE), 1)),
  Min = sapply(F23_bacteria[key_vars], function(x) round(min(x, na.rm = TRUE), 1)),
  Q25 = sapply(F23_bacteria[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 1)),
  Median = sapply(F23_bacteria[key_vars], function(x) round(median(x, na.rm = TRUE), 1)),
  Q75 = sapply(F23_bacteria[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 1)),
  Max = sapply(F23_bacteria[key_vars], function(x) round(max(x, na.rm = TRUE), 1))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(F23_bacteria$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(F23_bacteria$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(F23_bacteria$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(F23_bacteria$LB_Batch), "\n")
cat("- Sample types:", paste(unique(F23_bacteria$SampleType), collapse = ", "), "\n")
cat("- Platforms:", paste(unique(F23_bacteria$Platform), collapse = ", "), "\n")

# Data Distribution Plots
cat("\n=== Data Distribution Plots ===\n")

# Create plots for each variable
p1 <- create_hist_density(F23_bacteria, "TotalCells", "Total Bacteria Distribution", "Total Bacteria")
p2 <- create_hist_density(F23_bacteria, "LIVECounts", "Live Bacteria Distribution", "Live Bacteria")
p3 <- create_hist_density(F23_bacteria, "TotalCells_corrected.m3", "Bacteria Concentration Distribution", "Bacteria concentration (cells/m³)")
p4 <- create_hist_density(F23_bacteria, "LiveRatio", "Live Bacteria Ratio Distribution", "Live Bacteria Ratio")
p5 <- create_hist_density(F23_bacteria, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p6 <- create_hist_density(F23_bacteria, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p7 <- create_hist_density(F23_bacteria, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p8 <- create_hist_density(F23_bacteria, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p9 <- create_hist_density(F23_bacteria, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p10 <- create_hist_density(F23_bacteria, "MeanMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) / (p9 + p10) + 
  plot_annotation(
    title = "FASMEE23 Bacteria Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )
```

```{r FASMEE23 bacteria exploratory visualizations, echo=FALSE}
# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")

bp1 <- create_boxplot(F23_bacteria, "TotalCells_corrected.m3", "Bacteria Concentration by Smoke Level")
bp2 <- create_boxplot(F23_bacteria, "LiveRatio", "Live Bacteria Ratio by Smoke Level")
bp3 <- create_boxplot(F23_bacteria, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp4 <- create_boxplot(F23_bacteria, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")

# Select numeric variables for correlation
cor_vars <- c("TotalCells", "LIVECounts", "TotalCells_corrected.m3", "LiveRatio", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MeanMCE")
cor_data <- F23_bacteria[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Bacteria vs PM2.5
sp1 <- ggplot(F23_bacteria, aes(x = MedianPM2.5_ug.m3, y = TotalCells_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Bacteria Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Bacteria Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Bacteria vs Carbon
sp2 <- ggplot(F23_bacteria, aes(x = ALL_carbon_mg.m3, y = TotalCells_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Bacteria Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Bacteria Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Live Ratio vs PM2.5
sp3 <- ggplot(F23_bacteria, aes(x = MedianPM2.5_ug.m3, y = LiveRatio, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Live Bacteria Ratio vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Live Bacteria Ratio") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Live vs Total Bacteria
sp4 <- ggplot(F23_bacteria, aes(x = TotalCells, y = LIVECounts, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Live vs Total Bacteria", 
       x = "Total Bacteria", y = "Live Bacteria") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1 + sp2) / (sp3 + sp4) +
  plot_annotation(title = "Key Variable Relationships")

# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- F23_bacteria %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_bacteria = mean(TotalCells_corrected.m3, na.rm = TRUE),
    sd_bacteria = sd(TotalCells_corrected.m3, na.rm = TRUE),
    mean_live_ratio = mean(LiveRatio, na.rm = TRUE),
    sd_live_ratio = sd(LiveRatio, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

## [8.1] GLMM model for FASMEE23 bacteria and DHARMa diagnostics for assessment
```{r FASMEE23 bacteria GLMM}
F23_Bacteria_smokelvl_model <- glmmTMB(TotalCells ~ SmokeLevel + offset(log(RepVolume_L)) + 
                           (1|Platform:SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = F23_bacteria, 
                         ziformula = ~ SmokeLevel, dispformula = ~ (1|Platform:SampleID))
summary(F23_Bacteria_smokelvl_model)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = F23_Bacteria_smokelvl_model, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = F23_bacteria$SmokeLevel) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(F23_Bacteria_smokelvl_model, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(F23_Bacteria_smokelvl_model, test = "Chisq")) 

# Calculate marginal means and contrasts
emm_log <- emmeans(F23_Bacteria_smokelvl_model, ~ SmokeLevel)
emm_response <- emmeans(F23_Bacteria_smokelvl_model, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
bacteria_smokelvl <- as.data.frame(emm_response) %>%
  mutate(
    Bacteria.m3 = ((response - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025,
    CI_lower = ((asymp.LCL - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025,
    CI_upper = ((asymp.UCL - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025
  )

print(bacteria_smokelvl)

```
#### Smoke level is not a significant predictor in this model, lets try smoke vs ambient

```{r}

FASMEE23_BacteriaCatModel <- glmmTMB(TotalCells ~ SampleType + offset(log(RepVolume_L)) + 
                           (1|SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = F23_bacteria, 
                         ziformula = ~ 1, dispformula = ~ (1|SampleID))
summary(FASMEE23_BacteriaCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_BacteriaCatModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = F23_bacteria$SampleType) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_BacteriaCatModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(FASMEE23_BacteriaCatModel, test = "Chisq")) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_BacteriaCatModel, ~ SampleType)
emm_response <- emmeans(FASMEE23_BacteriaCatModel, ~ SampleType, type = "response")
F23_bact_rate_ratios <- pairs(emm_response, reverse = T)

# Create prediction dataframe
bacteria_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Bacteria.m3 = ((response - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025,
    CI_lower = ((asymp.LCL - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025,
    CI_upper = ((asymp.UCL - (F23_FB_bacteria_stat$Median_TotBacteria.fov + F23_LB_bacteria_stat$median_TotBacteria.fov)) * 7225)/0.025
  )

```


## [8.2] Predictions from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria Marginal means, echo=FALSE}
print(bacteria_cat)
```

## [8.3] Ratios from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria contrast-results, echo=FALSE}
print(summary(F23_bact_rate_ratios, infer = T))
```

## [8.4] Plotting FASMEE23 bacteria model results with raw data as points/violin plots
```{r FASMEE23 bacteria Plotting model results, fig.cap="FASMEE23 Bacteria GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
F23_bacteria_contrast <- as.data.frame(F23_bact_rate_ratios)
F23_bacteria_sig_bars <- data.frame(
  x = c(1),
  xend = c(2),
  pair = c("Smoke / Ambient"),
  stringsAsFactors = FALSE
)

F23_bacteria_sig_bars$p_value <- sapply(F23_bacteria_sig_bars$pair, function(pair) {
  idx <- which(F23_bacteria_contrast$contrast == pair)
  if(length(idx) > 0) {
    return(F23_bacteria_contrast$p.value[idx])
  } else {
    return(1.0) 
  }
})

F23_bacteria_max_height <- max(bacteria_cat$CI_upper, na.rm = TRUE)
F23_bacteria_sig_bars$y <- F23_bacteria_max_height  * c(5.5)
F23_bacteria_sig_bars$annotation <- ifelse(
  is.na(F23_bacteria_sig_bars$p_value),
  'plain("")',
  paste0("italic(p)==", signif(F23_bacteria_sig_bars$p_value, 2))
)


F23_bacteria_plt <- ggplot(bacteria_cat, aes(x = SampleType, y = Bacteria.m3)) + 
  geom_point(data = F23_bacteria, 
             aes(x = SampleType, y = TotalCells_corrected.m3, color = "Raw Data"), 
             position = position_jitter(width = 0.15), 
             alpha = 0.3, size = 1) +
  geom_violin(data = F23_bacteria, 
            aes(x = SampleType, y = TotalCells_corrected.m3), 
            fill = NA, color = "gold3", width = 0.8) +  
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "blue2", alpha = 0.6, linewidth = 0.8) +
  geom_point(aes(y = Bacteria.m3, color = "Model Output"), size = 1.5) +
  geom_segment(data = F23_bacteria_sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = F23_bacteria_sig_bars,
               aes(x = x, xend = x, y = y, yend = y - F23_bacteria_max_height  * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_segment(data = F23_bacteria_sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - F23_bacteria_max_height  * 0.09),
               linewidth = 0.5, color = "blue2") +
  geom_text(data = F23_bacteria_sig_bars,
            aes(x = (x + xend) / 2, y = y + F23_bacteria_max_height * 0.02, label = annotation),
            size = 2, vjust = 0, parse = TRUE) +
  geom_segment(aes(x = as.numeric(SampleType), 
                   xend = as.numeric(SampleType) + 0.2,  
                   y = Bacteria.m3, 
                   yend = Bacteria.m3 + 600000),  
               linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_label(data = bacteria_cat, 
            aes(label = paste0("hat(mu) == ", 
                              round(Bacteria.m3/10^floor(log10(Bacteria.m3)), 2), 
                              " %*% 10^", floor(log10(Bacteria.m3))),
                x = as.numeric(SampleType) + 0.2,  
                y = Bacteria.m3 + 600000),
            hjust = 0, size = 2, parse = TRUE,
            fill = "white", color = "black", alpha = 0.3,
            label.padding = unit(0.25, "lines"),
            label.r = unit(0.15, "lines")) +
  scale_color_manual(values = c("Model Output" = "blue2", "Raw Data" = "gold3")) +
  guides(color = guide_legend(override.aes = list(shape = 15, size = 4, alpha = 1))) +
  labs(x = "Sample Type", 
       y = expression("Subalpine forest: bacteria "~m^{-3}),
       color = "") +
  theme_bw() +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        panel.grid = element_blank(),
        legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.key = element_rect(fill = "white", color = NA)) +
  coord_cartesian(ylim = c(0, max(F23_bacteria$TotalCells_corrected.m3 * 0.4)))

print(F23_bacteria_plt)

ggsave('Output/Output_figs/FASMEE23/F23_SmokeVsAmbient_BacteriaModel.png', F23_bacteria_plt, dpi = 600, width = 5, height = 4, units = "in")

```

```{r, plotting for manuscript}

F23_spore_plt <- F23_spore_plt +
  labs(x = element_blank()) +
  theme(legend.position = "none")


k4_spore_plt <- k4_spore_plt +
  labs(x = element_blank()) 

F23_bacteria_plt <- F23_bacteria_plt +
  theme(legend.position = "none")

K4_F23_amb_vs_smoke <- (k4_spore_plt / F23_spore_plt / F23_bacteria_plt)

print(K4_F23_amb_vs_smoke)

ggsave('Output/Output_figs/K4_F23_SmokeVsAmbient_Models.png', K4_F23_amb_vs_smoke, dpi = 600, width = 4, height = 9, units = "in")
```

#### Live bacteria Ambient vs. Smoke
```{r}

FASMEE23_LiveBac_CatModel <- glmmTMB(LIVECounts ~ SampleType + offset(log(RepVolume_L)) + 
                           (1|SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = F23_bacteria, 
                         ziformula = ~ 1, dispformula = ~ (1|SampleID))
summary(FASMEE23_LiveBac_CatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_LiveBac_CatModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = F23_bacteria$SampleType) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_LiveBac_CatModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(FASMEE23_LiveBac_CatModel, test = "Chisq")) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_LiveBac_CatModel, ~ SampleType)
emm_response <- emmeans(FASMEE23_LiveBac_CatModel, ~ SampleType, type = "response")
F23_Livebact_rate_ratios <- pairs(emm_response, reverse = T)

# Create prediction dataframe
LiveBact_cat <- as.data.frame(emm_response) %>% 
  mutate(
    LiveBacteria.m3 = ((response - (F23_FB_bacteria_stat$Median_LiveBacteria.fov + F23_LB_bacteria_stat$median_LiveBacteria.fov)) * 7225)/0.025,
    CI_lower = ((asymp.LCL - (F23_FB_bacteria_stat$Median_LiveBacteria.fov + F23_LB_bacteria_stat$median_LiveBacteria.fov)) * 7225)/0.025,
    CI_upper = ((asymp.UCL - (F23_FB_bacteria_stat$Median_LiveBacteria.fov + F23_LB_bacteria_stat$median_LiveBacteria.fov)) * 7225)/0.025
  )

print(summary(F23_Livebact_rate_ratios, infer = T))
cat("\n")
print(LiveBact_cat)

```
#### Dead bacteria Ambient vs. Smoke
```{r}

FASMEE23_DeadBac_CatModel <- glmmTMB(DEADCounts ~ SampleType + offset(log(RepVolume_L)) + 
                           (1|SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = F23_bacteria, 
                         ziformula = ~ 1, dispformula = ~ 1) 
summary(FASMEE23_DeadBac_CatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_DeadBac_CatModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = F23_bacteria$SampleType) # Residuals by group
plotResiduals(res)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_DeadBac_CatModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(FASMEE23_DeadBac_CatModel, test = "Chisq")) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_DeadBac_CatModel, ~ SampleType)
emm_response <- emmeans(FASMEE23_DeadBac_CatModel, ~ SampleType, type = "response")
F23_Deadbact_rate_ratios <- pairs(emm_response, reverse = T)

# Create prediction dataframe
DeadBact_cat <- as.data.frame(emm_response) %>% 
  mutate(
    DeadBacteria.m3 = ((response - (F23_FB_bacteria_stat$Median_DeadBacteria.fov + F23_LB_bacteria_stat$median_DeadBacteria.fov)) * 7225)/0.025,
    CI_lower = ((asymp.LCL - (F23_FB_bacteria_stat$Median_DeadBacteria.fov + F23_LB_bacteria_stat$median_DeadBacteria.fov)) * 7225)/0.025,
    CI_upper = ((asymp.UCL - (F23_FB_bacteria_stat$Median_DeadBacteria.fov + F23_LB_bacteria_stat$median_DeadBacteria.fov)) * 7225)/0.025
  )

print(summary(F23_Deadbact_rate_ratios, infer = T))
cat("\n")
print(DeadBact_cat)

```
# [9] Modeling for research question #3: FASMEE23 live bacteria ratio in ambient vs. smoke conditions
```{r FASMEE23 Live bacteria ratio}


#cbind(DEADCounts, TotalCells - DEADCounts)
#cbind(LIVECounts, TotalCells - LIVECounts)


FASMEE23_LiveRatioModel <- glmmTMB(cbind(LIVECounts, TotalCells - LIVECounts) ~ SampleType +
                          (1|SampleID) + (1|LB_Batch),
                          family = betabinomial(link = "logit"), 
                          data = F23_bacteria, dispformula = ~(1|SampleID))
summary(FASMEE23_LiveRatioModel)


res <- simulateResiduals(fittedModel = FASMEE23_LiveRatioModel, plot = FALSE)
plotQQunif(res)
testCategorical(res, catPred = F23_bacteria$SampleType) # Residuals by group
plotResiduals(res)
plotResiduals(res, F23_bacteria$MeanMR)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_LiveRatioModel, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(FASMEE23_LiveRatioModel, test = "Chisq")) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_LiveRatioModel, ~ SampleType)
emm_response <- emmeans(FASMEE23_LiveRatioModel, ~ SampleType, type = "response")
F23_Livebact_rate_ratios <- pairs(emm_response, reverse = T)

```

# [10] Modeling for research question #2: Environmental drivers of Konza4 smoke spore abundance


## [10.1] GLMM model for Konza4 smoke spore abundance vs. environmental drivers and DHARMa diagnostics for assessment
```{r Konza4 test smoke spore GLMM}

Konza4_Smoke_model <- glmmTMB(TotalSpores ~ p5.0um.dl + MeanMR + MeanRH + offset(log(RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = Konza4_smoke_spores)
summary(Konza4_Smoke_model)


res <- simulateResiduals(fittedModel = Konza4_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)

simOutput_cond <- simulateResiduals(Konza4_Smoke_model, re.form = NULL)
testDispersion(simOutput_cond)


```

```{r Konza4 smoke spore GLMM}
Konza4_Smoke_model <- glmmTMB(TotalSpores ~ MedianPM2.5_ug.m3 + MeanMR + MeanRH + offset(log(RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = Konza4_smoke_spores)
summary(Konza4_Smoke_model)

res <- simulateResiduals(fittedModel = Konza4_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)

simOutput_cond <- simulateResiduals(Konza4_Smoke_model, re.form = NULL)
testDispersion(simOutput_cond)


```
## [10.2] Plotting Konza4 smoke spore model vs. environmental drivers with raw data as points
```{r Konza4 smoke spore model plotting, fig.width=8, fig.height=5, fig.cap="Konza4 Smoke Spore Abundance vs. PM2.5 and Moisture Ratio"}
# PM2.5 predictions
pm_pred <- emmeans(Konza4_Smoke_model, ~ MedianPM2.5_ug.m3, 
                  at = list(MedianPM2.5_ug.m3 = seq(min(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   max(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   length.out = 100)),
                  type = "response")
pm_df <- as.data.frame(pm_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Moisture Ratio predictions
mr_pred <- emmeans(Konza4_Smoke_model, ~ MeanMR,
                  at = list(MeanMR = seq(min(Konza4_smoke_spores$MeanMR),
                                       max(Konza4_smoke_spores$MeanMR),
                                       length.out = 100)),
                  type = "response")
mr_df <- as.data.frame(mr_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# PM2.5 plot
p_pm <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3),
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_ribbon(data = pm_df,
              aes(x = MedianPM2.5_ug.m3, ymin = CI_lower, ymax = CI_upper),
              fill = "#0072B2", alpha = 0.2) +
  geom_line(data = pm_df,
            aes(x = MedianPM2.5_ug.m3, y = Spores.m3),
            color = "#0072B2", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~("μg"~m^{-3})),
       y = expression(Spores~m^{-3}),
       title = "PM2.5") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Moisture Ratio plot
p_mr <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MeanMR, y = TotalSpores_corrected.m3),
             alpha = 0.5, size = 2, color = "#D55E00") +
  geom_ribbon(data = mr_df,
              aes(x = MeanMR, ymin = CI_lower, ymax = CI_upper),
              fill = "#D55E00", alpha = 0.2) +
  geom_line(data = mr_df,
            aes(x = MeanMR, y = Spores.m3),
            color = "#D55E00", linewidth = 1.1) +
  labs(x = "Moisture Ratio",
       y = expression(Spores~m^{-3}),
       title = "Moisture Ratio") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Combine with patchwork
(p_pm | p_mr) + plot_annotation(title = "Konza4 Smoke Spore Abundance vs. Environmental Variables")
```

# [11] Modeling for research question #2: Environmental drivers of FASMEE23 smoke spore abundance

## [11.1] GLMM model for FASMEE23 smoke spore abundance vs. environmental drivers and DHARMa diagnostics for assessment

### FASMEE23 smoke spore abundance vs. environmental drivers model shows poor fit and no significant effects
```{r FASMEE23 smoke spore GLMM}

FASMEE23_Smoke_model <- glmmTMB(TotalSpores ~ MedianPM2.5_ug.m3 + offset(log(RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = FASMEE23_smoke_spores, 
                       ziformula = ~-1, dispformula = ~(1|SampleID))
summary(FASMEE23_Smoke_model)

res <- simulateResiduals(fittedModel = FASMEE23_Smoke_model, plot = F)
plotResiduals(res)
plotResiduals(res, FASMEE23_smoke_spores$MeanRH)
plotQQunif(res)

testZeroInflation(res)

simOutput_cond <- simulateResiduals(FASMEE23_Smoke_model, re.form = NULL)
testDispersion(simOutput_cond) 
```
# [12] Modeling for research question #2: Environmental drivers of FASMEE23 smoke bacteria abundance

## [12.1] GLMM model for FASMEE23 smoke bacteria abundance vs. environmental drivers and DHARMa diagnostics for assessment

### FASMEE23 smoke bacteria abundance vs. environmental drivers model shows poor fit and no significant effects
```{r FASMEE23 smoke bacteria GLMM}

FASMEE23_Smoke_model <- glmmTMB(TotalCells ~ log(p1.0um.dl) + offset(log(RepVolume_L)) + 
                          (1|SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_smoke_bacteria, 
                        ziformula = ~1, dispformula = ~(1|SampleID))
summary(FASMEE23_Smoke_model)

res <- simulateResiduals(fittedModel = FASMEE23_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_Smoke_model, re.form = NULL)
testDispersion(simOutput_cond) 
```
# [13] Site comparison data summary
```{r Konza4 and FASMEE23 site comparison data}

fasmee23_k4_spores <- bind_rows(Konza4_spores_pa_C, FASMEE23_spores_pa_C) %>%
  mutate(
    Project_SampleType = paste(Project, SampleType, sep = "_"),
    Project_SampleType = factor(Project_SampleType, levels = c("Konza4_Ambient", "Konza4_Smoke", "FASMEE23_Ambient", "FASMEE23_Smoke")),
    Project_platform = if_else(Project == "FASMEE23", paste0(Project, "_", Platform), paste0("Konza4", "_", "Blue"))
    ) 

```

```{r Konza4 and FASMEE23 site comparison plot, echo=FALSE}
# Site Comparison: FASMEE23 vs Konza4
cat("\n=== Site Comparison: FASMEE23 vs Konza4 ===\n")

df_spores <- fasmee23_k4_spores

# Descriptive statistics by site and sample type
cat("\nDescriptive statistics by site and sample type:\n")
spore_stats <- df_spores %>%
  group_by(Project, SampleType) %>%
  summarise(
    n = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_rh = mean(MeanRH, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))
print(spore_stats)

# Boxplot: Spore concentration by site and sample type
cat("\nBoxplot: Spore concentration by site and sample type\n")
p1 <- ggplot(df_spores, aes(x = Project_SampleType, y = TotalSpores_corrected.m3, fill = Project)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  labs(title = "Spore Concentration by Site and Sample Type", x = "Site & Sample Type", y = "Spores (cells/m³)") +
  scale_fill_manual(values = c("Konza4" = "#56B4E9", "FASMEE23" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.text.x = element_text(angle = 30, hjust = 1))
print(p1)

```
# [14] Modeling for research question #4: Site comparison of Konza4 and FASMEE23
## [14.1] GLMM model for Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison GLMM}

fasmee23_k4_spores_model <- glmmTMB(TotalSpores ~ Project_SampleType + offset(log(RepVolume_L)) + 
                                    (1|Project_platform:SampleID) + (1|LB_Batch),
                                  family=nbinom2(link="log"), data = fasmee23_k4_spores, 
                                  dispformula = ~(1|Project_platform:SampleID))
summary(fasmee23_k4_spores_model)

res <- simulateResiduals(fittedModel = fasmee23_k4_spores_model, plot = F)
plotResiduals(res)
plotQQunif(res)
testCategorical(res, catPred = fasmee23_k4_spores$Project_SampleType)
testZeroInflation(res)
simOutput_cond <- simulateResiduals(FASMEE23_Smoke_model, re.form = NULL)
testDispersion(simOutput_cond) 

print(drop1(fasmee23_k4_spores_model, test = "Chisq"))
```
## [14.2] Marginal means and contrasts for Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison marginal means and contrasts}
# Calculate marginal means and contrasts
emm_log <- emmeans(fasmee23_k4_spores_model, ~ Project_SampleType)
emm_response <- emmeans(fasmee23_k4_spores_model, ~ Project_SampleType, type = "response")

rate_ratios <- pairs(emm_log, type = "response", reverse = T)

site_comparison_cat <- as.data.frame(emm_response) %>%
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

contrast_df <- as.data.frame(rate_ratios)

print(summary(rate_ratios, infer = T))
print(site_comparison_cat)
```
## [14.3] Plotting Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison plotting}
sig_bars <- data.frame(
  x = c(1, 3, 1, 2),
  xend = c(2, 4, 3, 4),
  pair = c("Konza4 Smoke vs Ambient", "FASMEE23 Smoke vs Ambient",
           "Ambient: FASMEE23 vs Konza4", "Smoke: FASMEE23 vs Konza4"),
  stringsAsFactors = FALSE
)
sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0)
  }
})
max_height <- max(site_comparison_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3, 3.5, 3.7)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)==", signif(sig_bars$p_value, 2))
)

# Helper function for scientific plotmath labels
sci_label <- function(x) {
  if (x == 0) return("0")
  expo <- floor(log10(abs(x)))
  base <- signif(x / 10^expo, 2)
  paste0("hat(mu) == ", base, " %*% 10^", expo)
}

p1 <- ggplot(site_comparison_cat, aes(x = Project_SampleType, y = Spores.m3)) +
  geom_point(data = fasmee23_k4_spores,
             aes(x = Project_SampleType, y = TotalSpores_corrected.m3),
             position = position_jitter(width = 0.1),
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = fasmee23_k4_spores,
              aes(x = Project_SampleType, y = TotalSpores_corrected.m3),
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = "Site and Sample Type",
       y = expression(Spores~m^{-3}),
       title = "Fungal Spore Abundance: Site Comparison") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("Konza4\nAmbient", "Konza4\nSmoke",
                              "FASMEE23\nAmbient", "FASMEE23\nSmoke")) +
  geom_label(
    data = site_comparison_cat,
    aes(
      x = as.numeric(Project_SampleType) + 0.16,
      y = Spores.m3,
      label = sapply(Spores.m3, sci_label)
    ),
    parse = TRUE,
    hjust = 0,
    vjust = -0.2,
    fill = "white",
    color = "black",
    label.size = 0.3,
    fontface = "plain",
    size = 2.5
  ) +
  coord_cartesian(ylim = c(0, max(fasmee23_k4_spores$TotalSpores_corrected.m3 * 0.5)))

print(p1)
```


# [15] Emission Factors

```{r Emission Factors}
calculate_ef_stats <- function(data, ef_columns) {
  results <- map_dfr(ef_columns, ~{
    data %>%
      summarise(
        column = .x,
        EF_value = mean(!!sym(.x), na.rm = TRUE),
        SE_value = sd(!!sym(.x), na.rm = TRUE) / sqrt(sum(!is.na(!!sym(.x)))),
        .groups = "drop"
      ) %>%
      mutate(
        Lower_CI = EF_value - 1.96 * SE_value,
        Upper_CI = EF_value + 1.96 * SE_value
      )
  })
  return(results)
}

# Calculate emission factors for all groups using already-imported data
# Konza4 bacteria
konza4_bacteria_ef_stats <- K4_bacteria %>%
  filter(!is.na(MeanMCE)) %>%
  mutate(Tot_bacteria_g.Mg = Total_bacteria.Mg*6.0e-14,
         Live_bacteria_g.Mg = Live_bacteria.Mg*6.0e-14) %>% #mass of bacteria ~6.0×10-14 g
  calculate_ef_stats(c("Total_bacteria.Mg", "Tot_bacteria_g.Mg", "Live_bacteria.Mg", "Live_bacteria_g.Mg"))

# Konza4 Spores
konza4_spores_ef_stats <- Konza4_smoke_spores %>%
  filter(!is.na(MeanMCE)) %>%
  mutate(spores_g.Mg = spores.Mg*3.3e-11) %>% # mass of a fungal spore ~ 3.3×10-11 g
  calculate_ef_stats(c("spores.Mg", "spores_g.Mg"))

# FASMEE23 Bacteria
fasmee23_bacteria_ef_stats <- FASMEE23_smoke_bacteria %>%
  filter(!is.na(MeanMCE)) %>%
  mutate(Tot_bacteria_g.Mg = Total_bacteria.Mg*6.0e-14,
         Live_bacteria_g.Mg = Live_bacteria.Mg*6.0e-14) %>% #mass of bacteria ~6.0×10-14 
  calculate_ef_stats(c("Total_bacteria.Mg", "Tot_bacteria_g.Mg", "Live_bacteria.Mg", "Live_bacteria_g.Mg"))

# FASMEE23 Spores
fasmee23_spores_ef_stats <- FASMEE23_smoke_spores %>%
  filter(!is.na(MeanMCE)) %>%
  mutate(spores_g.Mg = spores.Mg*3.3e-11) %>% # mass of a fungal spore ~ 3.3×10-11 g
  calculate_ef_stats(c("spores.Mg", "spores_g.Mg"))

ef_df <- bind_rows(
  konza4_bacteria_ef_stats %>% mutate(Study = "Konza4"),
  konza4_spores_ef_stats %>% mutate(Study = "Konza4"),
  fasmee23_bacteria_ef_stats %>% mutate(Study = "FASMEE23"),
  fasmee23_spores_ef_stats %>% mutate(Study = "FASMEE23")
) %>%
  select(Study, Measurment = column, EF_value, SE_value, Lower_CI, Upper_CI) %>%
  arrange(Study, Measurment)

print("Final EF DataFrame:")
print(ef_df)

```
## EF plotting

```{r}
# Split data for plots
# konza4_df <- ef_df %>% filter(Study == "Konza4")
# fasmee_df <- ef_df %>% filter(Study == "FASMEE23")
# 
# # Konza4 plot
# p_konza4 <- ggplot(konza4_df, aes(x = Measurment, y = EF_value, fill = Measurment)) +
#   geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.18, linewidth = 1.1) +
#   labs(
#     x = NULL,
#     y = NULL,
#     title = "Kansas Tallgrass Prairie"
#   ) +
#   theme_bw(base_size = 12) +
#   theme(
#     legend.position = "none",
#     axis.text.x = element_text(angle = 20, hjust = 1, size = 12, face = "bold"),
#     axis.title.y = element_text(size = 12, face = "bold"),
#     plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
#   )
# 
# # FASMEE23 plot
# p_fasmee <- ggplot(fasmee_df, aes(x = Group, y = EF, fill = Group)) +
#   geom_col_pattern(
#    width = 0.7, 
#    color = "black",
#    alpha = 0.30,
#    fill = "white",              
#    pattern = c("stripe", "crosshatch"),
#    pattern_angle = c(35, 0),
#    pattern_density = c(0.1, 0.15),
#    pattern_spacing = c(0.05, 0.05),
#    pattern_fill = "black",
#    pattern_alpha = 0.5
#   ) +
#   geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.18, linewidth = 1.1) +
#   labs(
#     x = NULL,
#     y = expression("Microbes "*Mg^-1),
#     title = "Utah Subalpine Forest"
#   ) +
#   theme_bw(base_size = 12) +
#   theme(
#     legend.position = "none",
#     axis.text.x = element_text(angle = 20, hjust = 1, size = 12, face = "bold"),
#     axis.title.y = element_text(size = 12, face = "bold"),
#     plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
#   )
# 
# # Combine with patchwork
# library(patchwork)
# 
# combined_plot <- (p_fasmee | p_konza4) + 
#   plot_annotation(
#     title = "Microbial Emission Factors from Wildland Fires",
#     caption = "Error bars: 95% CI"
#   )
# 
# print(combined_plot)

#ggsave("Output/Output_figs/K4_FASMEE23_EF.png", combined_plot, width = 6.5, height = 4, dpi = 600)
```
##Scaling EFs up to Monroe Canyon and Four County Fires
```{r}

four_county_ha = 132933/2.471 # grassland/cropland hectares according to https://www.nass.usda.gov/Research_and_Science/Disaster-Analysis/2021/Fire_2021_Kansas/Kansas_Report_12-23-21.pdf

monroe_ha = 36637/2.471 # hectares on 7/30/2025 according to NIFC (obtained from WatchDuty)

monroe_fuel_Mg.ha = 95.81 # average of Kobziar et al. 2024 BEF paper, as outlined in Hudak et al. 2024
four_country_fuel_Mg.ha = 5.415

bacteria_mass_g = 6e-14 # Kobziar et al. 2024 
spore_mass_g = 3.3e-11 # Sesartic et al. 

monroe_fungi_spores.Mg = 2.493874e13
monroe_spores_g.Mg = 8.229783e+02
monroe_bacteria.Mg = 1.084549e14
monroe_bacteria_g.Mg = 6.507296e+00

four_county_bacteria.Mg = 8.860138e11
four_county_bacteria_g.Mg = 5.316083e-02
four_county_fungi_spores.Mg = 1.267984e12	
four_county_spores_g.Mg = 4.184347e+01

  
monroe_totalfuel_Mg = monroe_fuel_Mg.ha * monroe_ha
four_county_totalfuel_Mg = four_country_fuel_Mg.ha * four_county_ha
  
monroe_fungi = monroe_fungi_spores.Mg * monroe_totalfuel_Mg
monroe_fungi_mass = monroe_spores_g.Mg * monroe_totalfuel_Mg
four_county_fungi = four_county_fungi_spores.Mg * four_county_totalfuel_Mg
four_county_fungi_mass = four_county_spores_g.Mg * four_county_totalfuel_Mg

monroe_bacteria = monroe_bacteria.Mg * monroe_totalfuel_Mg
monroe_bacteria_mass = monroe_bacteria_g.Mg * monroe_totalfuel_Mg
four_county_bacteria = four_county_bacteria.Mg * four_county_totalfuel_Mg
four_county_bacteria_mass = four_county_bacteria_g.Mg * four_county_totalfuel_Mg

monroe_total_crobe = monroe_fungi + monroe_bacteria
monroe_total_crobe_mass = monroe_fungi_mass + monroe_bacteria_mass
four_total_crobe = four_county_fungi + four_county_bacteria
four_total_crobe_mass = four_county_fungi_mass + four_county_bacteria_mass

cat("Monroe total spores\n")
print(monroe_fungi)
cat("\nMonroe total spore mass\n")
print(monroe_fungi_mass)
cat("\nMonroe total bacteria\n")
print(monroe_bacteria)
cat("\nMonroe total bacteria mass\n")
print(monroe_bacteria_mass)
cat("\nFour County total spores\n")
print(four_county_fungi)
cat("\nFour County total spore mass\n")
print(four_county_fungi_mass)
cat("\nFour Country total bacteria\n")
print(four_county_bacteria)
cat("\nFour Country total bacteria mass\n")
print(four_county_bacteria_mass)
cat("\nMonroe total microbes (bacteria + fungal spores)\n")
print(monroe_total_crobe)
cat("\nMonroe total microbe mass (bacteria + fungal spores)\n")
print(monroe_total_crobe_mass)
cat("\nFour County total microbes (bacteria + fungal spores)\n")
print(four_total_crobe)
cat("\nFour County total microbe mass (bacteria + fungal spores)\n")
print(four_total_crobe_mass)

```

## Modeling for MCE (incomplete!!!)
```{r}
# Konza4 Spores
konza4_spores_ef <- Konza4_spores_pa_C %>%
  filter(SampleType == "Smoke" & !is.na(MeanMCE))

k4_spore_mce <- glmmTMB(spores.mg ~ MeanMCE + (1|SampleID),
                        family = tweedie(link = "log"), data = konza4_spores_ef)
summary(k4_spore_mce)

res <- simulateResiduals(fittedModel = k4_spore_mce, plot = F)
plotResiduals(res)
plotResiduals(res, konza4_spores_ef$MeanMCE)
plotQQunif(res)

testZeroInflation(res)

simOutput_cond <- simulateResiduals(k4_spore_mce, re.form = NULL)
testDispersion(simOutput_cond)

# Konza4 Spores by sample

konza4_spores_ef_sample <- konza4_spores_ef %>%
  group_by(SampleID) %>%
  summarise(
    spores.Mg = mean(spores.Mg),
    MeanMCE = mean(MeanMCE)
  )

k4_spore_mce_sample <- glm(spores.Mg ~ MeanMCE,
                        family = Gamma(link = "log"), 
                    data = konza4_spores_ef_sample)
summary(k4_spore_mce_sample)

# res <- simulateResiduals(fittedModel = k4_spore_mce_sample, plot = F)
# plotResiduals(res)
# plotResiduals(res, konza4_spores_ef$MeanMCE)
# plotQQunif(res)
# 
# testZeroInflation(res)
# 
# simOutput_cond <- simulateResiduals(k4_spore_mce, re.form = NULL)
# testDispersion(simOutput_cond)




# FASMEE23 Bacteria
fasmee23_bacteria_ef <- F23_BlueBacteria %>%
  filter(SampleType == "Smoke" & !is.na(MeanMCE)) 

# FASMEE23 Spores
fasmee23_spores_ef <- F23_BlueSpores %>%
  filter(SampleType == "Smoke" & !is.na(MeanMCE)) 


```

# [16] 18s qPCR spore enumeration comparison

```{r}

summary(F23_qPCR$DNA18s_copies_FBcorr.m3, na.rm = T)

qPCRvsCell <- left_join(F23_qPCR, FASMEE23_spores_pa_C, by = c("SampleID_short")) %>%
  filter(!is.na(TotalSpores))

qPCRvsCell_m <- glmmTMB(TotalSpores.filter ~ DNA18s_copies.sample + offset(log(RepVolume_L)) + 
                          (1|Platform:SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = qPCRvsCell, 
                        ziformula = ~-1, dispformula = ~ (1|Platform:SampleID))
summary(qPCRvsCell_m)

ggplot(qPCRvsCell, aes(x = DNA18s_copies.sample, y = TotalSpores.filter)) +
  geom_point() +
  labs(x = "DNA18s copies per sample", 
       y = "Total Spores (filtered)") +
  theme_minimal()
```

# [17] Emission rates (fluxes)
```{r}
k_turb0 = 10 ##(m2/s) -- Leads to the same range of values as burrows and Sesartic 
k_turb1 = 100 ##(m2/second) --  
```
## Konza 4 spores and bacteria
```{r}
# Konza 4 spores

Konza4_spores_pa_C <- Konza4_spores_pa_C %>%
  mutate(   ##Calculating Flux using method provided in Jacob (1999) and Sesartic and Dallafior (2011)                            
    deltaT0_s = (((MeanAGL)^2)/(2*k_turb0)),
    deltaT1_s = (((MeanAGL)^2)/(2*k_turb1)),
    flux0_spores.m2sec = ((TotalSpores_corrected.m3*MeanAGL)/(deltaT0_s)),
    flux1_spores.m2sec = ((TotalSpores_corrected.m3*MeanAGL)/(deltaT1_s)),
    flux_spores.m2sec = if_else(SampleType == "Ambient", 
                               ((TotalSpores_corrected.m3*MeanAGL)/(deltaT0_s)), 
                               ((TotalSpores_corrected.m3*MeanAGL)/(deltaT1_s)))
    )

K4_spore_flux_sample <- Konza4_spores_pa_C %>%
  group_by(Sample, SampleType) %>% 
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_spores.m2sec = mean(flux0_spores.m2sec),
    sd_flux0_spores.m2sec = sd(flux0_spores.m2sec),
    mean_flux1_spores.m2sec = mean(flux1_spores.m2sec),
    sd_flux1_spores.m2sec = sd(flux1_spores.m2sec),
    mean_flux_spores.m2sec = mean(flux_spores.m2sec),
    sd_flux_spores.m2sec = sd(flux_spores.m2sec)  
  )

cat("K4 spore flux by sample\n")
print(tibble(K4_spore_flux_sample))

K4_spore_flux_SampleType <- Konza4_spores_pa_C %>%
  group_by(SampleType) %>%
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_spores.m2sec = mean(flux0_spores.m2sec),
    sd_flux0_spores.m2sec = sd(flux0_spores.m2sec),
    mean_flux1_spores.m2sec = mean(flux1_spores.m2sec),
    sd_flux1_spores.m2sec = sd(flux1_spores.m2sec),
    mean_flux_spores.m2sec = mean(flux_spores.m2sec),
    sd_flux_spores.m2sec = sd(flux_spores.m2sec)  
  )

cat("\nK4 spore flux by sample type\n")
print(tibble(K4_spore_flux_SampleType))

# Konza 4 bacteria

K4_bacteria <- K4_bacteria %>%
  mutate(   ##Calculating Flux using method provided in Jacob (1999) and Sesartic and Dallafior (2011)                            
    deltaT0_s = (((MeanAGL)^2)/(2*k_turb0)),
    deltaT1_s = (((MeanAGL)^2)/(2*k_turb1)),
    flux0_cells.m2sec = ((TotalCells_corrected.m3*MeanAGL)/(deltaT0_s)),
    flux1_cells.m2sec = ((TotalCells_corrected.m3*MeanAGL)/(deltaT1_s)),
    flux_cells.m2sec = if_else(SampleType == "Ambient", 
                               ((TotalCells_corrected.m3*MeanAGL)/(deltaT0_s)), 
                               ((TotalCells_corrected.m3*MeanAGL)/(deltaT1_s)))
    )    

K4_bacteria_flux_sample <- K4_bacteria %>%
group_by(Sample, SampleType) %>% 
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_cells.m2sec = mean(flux0_cells.m2sec),
    sd_flux0_cells.m2sec = sd(flux0_cells.m2sec),
    mean_flux1_cells.m2sec = mean(flux1_cells.m2sec),
    sd_flux1_cells.m2sec = sd(flux1_cells.m2sec),
    mean_flux_cells.m2sec = mean(flux_cells.m2sec),
    sd_flux_cells.m2sec = sd(flux_cells.m2sec),
  )

cat("K4 bacteria flux by sample\n")
print(tibble(K4_bacteria_flux_sample))

K4_bacteria_flux_SampleType <- K4_bacteria %>%
  group_by(SampleType) %>%
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_cells.m2sec = mean(flux0_cells.m2sec),
    sd_flux0_cells.m2sec = sd(flux0_cells.m2sec),
    mean_flux1_cells.m2sec = mean(flux1_cells.m2sec),
    sd_flux1_cells.m2sec = sd(flux1_cells.m2sec),
    mean_flux_cells.m2sec = mean(flux_cells.m2sec),
    sd_flux_cells.m2sec = sd(flux_cells.m2sec),
  )

cat("\nK4 bacteria flux by sample type\n")
print(tibble(K4_bacteria_flux_SampleType))


```


## FASMEE23 spores and bacteria

```{r}
# F23 spores

F23_BlueSpores <- F23_BlueSpores %>%
  mutate(   ##Calculating Flux using method provided in Jacob (1999) and Sesartic and Dallafior (2011)                            
    deltaT0_s = (((MeanAGL)^2)/(2*k_turb0)),
    deltaT1_s = (((MeanAGL)^2)/(2*k_turb1)),
    flux0_spores.m2sec = ((TotalSpores_corrected.m3*MeanAGL)/(deltaT0_s)),
    flux1_spores.m2sec = ((TotalSpores_corrected.m3*MeanAGL)/(deltaT1_s)),
    flux_spores.m2sec = if_else(SampleType == "Ambient", 
                               ((TotalSpores_corrected.m3*MeanAGL)/(deltaT0_s)), 
                               ((TotalSpores_corrected.m3*MeanAGL)/(deltaT1_s)))
    )

F23_BlueSpores_flux_sample <- F23_BlueSpores %>%
group_by(SampleID_short, SampleType) %>%
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_spores.m2sec = mean(flux0_spores.m2sec),
    sd_flux0_spores.m2sec = sd(flux0_spores.m2sec),
    mean_flux1_spores.m2sec = mean(flux1_spores.m2sec),
    sd_flux1_spores.m2sec = sd(flux1_spores.m2sec),
    mean_flux_spores.m2sec = mean(flux_spores.m2sec),
    sd_flux_spores.m2sec = sd(flux_spores.m2sec)  
  ) %>% arrange(as.numeric(str_extract(SampleID_short, "\\d+")))

cat("F23 spore flux by sample\n")
print(tibble(F23_BlueSpores_flux_sample))

F23_BlueSpores_flux_SampleType <- F23_BlueSpores %>%
  group_by(SampleType) %>%
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_spores.m2sec = mean(flux0_spores.m2sec),
    sd_flux0_spores.m2sec = sd(flux0_spores.m2sec),
    mean_flux1_spores.m2sec = mean(flux1_spores.m2sec),
    sd_flux1_spores.m2sec = sd(flux1_spores.m2sec),
    mean_flux_spores.m2sec = mean(flux_spores.m2sec),
    sd_flux_spores.m2sec = sd(flux_spores.m2sec)  
  )

cat("\nF23 spore flux by sample type\n")
print(tibble(F23_BlueSpores_flux_SampleType))

# F23 bacteria

F23_BlueBacteria <- F23_BlueBacteria %>%
  mutate(   ##Calculating Flux using method provided in Jacob (1999) and Sesartic and Dallafior (2011)                            
    deltaT0_s = (((MeanAGL)^2)/(2*k_turb0)),
    deltaT1_s = (((MeanAGL)^2)/(2*k_turb1)),
    flux0_cells.m2sec = ((TotalCells_corrected.m3*MeanAGL)/(deltaT0_s)),
    flux1_cells.m2sec = ((TotalCells_corrected.m3*MeanAGL)/(deltaT1_s)),
    flux_cells.m2sec = if_else(SampleType == "Ambient", 
                               ((TotalCells_corrected.m3*MeanAGL)/(deltaT0_s)), 
                               ((TotalCells_corrected.m3*MeanAGL)/(deltaT1_s)))
    )    

F23_BlueBacteria_flux_sample <- F23_BlueBacteria %>%
group_by(SampleID_short, SampleType) %>% 
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_cells.m2sec = mean(flux0_cells.m2sec),
    sd_flux0_cells.m2sec = sd(flux0_cells.m2sec),
    mean_flux1_cells.m2sec = mean(flux1_cells.m2sec),
    sd_flux1_cells.m2sec = sd(flux1_cells.m2sec),
    mean_flux_cells.m2sec = mean(flux_cells.m2sec),
    sd_flux_cells.m2sec = sd(flux_cells.m2sec),
  ) %>% arrange(as.numeric(str_extract(SampleID_short, "\\d+")))

cat("F23 bacteria flux by sample\n")
print(tibble(F23_BlueBacteria_flux_sample))

F23_BlueBacteria_flux_SampleType <- F23_BlueBacteria %>%
  group_by(SampleType) %>%
  summarise(
    meanAGL = mean(MeanAGL),
    mean_flux0_cells.m2sec = mean(flux0_cells.m2sec),
    sd_flux0_cells.m2sec = sd(flux0_cells.m2sec),
    mean_flux1_cells.m2sec = mean(flux1_cells.m2sec),
    sd_flux1_cells.m2sec = sd(flux1_cells.m2sec),
    mean_flux_cells.m2sec = mean(flux_cells.m2sec),
    sd_flux_cells.m2sec = sd(flux_cells.m2sec),
  ) 

cat("\nF23 bacteria flux by sample type\n")
print(tibble(F23_BlueBacteria_flux_SampleType))


```



