---
title: "Smoke Bioaerosol Concentration and Emission Mixed Models"
format:
  html:
    code-fold: true
    df-print: paged
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    smooth-scroll: true
    fig-cap-location: bottom
    embed-resources: true
    code-summary: "Show code"
    code-tools: true
    margin-left: 2cm
    margin-right: 2cm
    code-block-width: 100%
  pdf:
    margin-left: 1.75cm
    margin-right: 1.75cm
    code-block-width: 100%
---

```{r setup, include=FALSE}
options(width = 80)  # Reduced from 110 to prevent text overflow
knitr::opts_chunk$set(
  dpi = 600,           # Higher DPI for better resolution
  fig.width = 6,      # Width in inches
  fig.height = 6,      # Height in inches
  out.width = "100%"   # How wide the image appears in the HTML
)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(emmeans)
library(glmmTMB)
library(DHARMa)
```

# Study Overview

## Site Description and Sampling Design

This study examined airborne microbes (fungi and bacteria) in ambient air and wildland fire smoke across two distinct ecosystems:

- **Utah subalpine** (FASMEE23): Subalpine forest fire regime
- **Kansas Tallgrass Prairie** (Konza4): Grassland fire regime

Samples were collected using drone-mounted equipment before fire ignition (ambient conditions) and during active burning (smoke conditions). Air and smoke samples were collected using filters and compensating pumps that drew specific volumes of air through the sample filters.

## Sample Collection Protocol

**Filter handling**: Before and after each sample collection, filters were placed and removed from sampling devices in situ using a mobile laboratory box to minimize air contamination and external influence on sample filters.

**Quality control**: Filter blanks were collected in situ to quantify contamination during filter handling and processing.

**Concurrent measurements**: Each drone flight simultaneously collected meteorological data, particulate matter (PM) concentrations, and carbon sensor data. *Note: Carbon sensor malfunctions resulted in incomplete carbon data for some samples.*

## Microbial Enumeration Methods

Microbes were quantified using epifluorescence microscopy after fluorescent staining:

**Counting protocol**:
- Magnification: 1000x
- Fields of view (FOV): 40 random locations per filter
- Filter area: ~226.98 mm²
- FOV area at 1000x: ~0.031 mm²
- Total possible FOVs per filter: 7,225


**Staining approaches**:
- **Bacteria**: Two-part viability stain allowing quantification of total bacteria, live bacteria, and dead bacteria
- **Fungi**: Single stain for spore enumeration (viability not distinguishable)

**Laboratory blanks**: Lab blanks were processed with each batch of staining solutions to account for contamination introduced during the staining process.

# Research Questions

1. **Microbial abundance across smoke conditions**: How do microbial concentrations in ambient air differ from concentrations in smoke of varying intensity?
   - Smoke intensity classified using PM<sub>2.5</sub> concentrations from onboard sensors

2. **Environmental drivers**: Do smoke-associated microbes correlate with meteorological variables or carbon concentrations?

3. **Bacterial viability**: Does the ratio of live to total bacteria differ between ambient and smoke conditions?

4. **Site comparison**: Are there differences in microbial abundances between study sites under both ambient and smoke conditions?

5. **Emission factors**: Using the carbon data, we can transform our microbe counts into the number of microbes emitted by a certain mass of fuel burning (e.g. #cells emitted per kilogram of biomass burning)---Known as emission factors (EFs)---Do EFs correlate with combustion efficiency (smoldering vs. flaming combustion)?

# Statistical Analysis Approach

## Generalized Linear Mixed Models (GLMMs)

**Response variable**: FOV count data (Research Questions 1-4)
**Offset**: log(volume sampled) to account for varying sample volumes

**Random effects structure**:
- Random intercept for SampleID (accounts for multiple FOVs from the same sample)
- Random intercept for Lab Blank batch (accounts for batch-specific contamination levels during sample processing)

*Note: Research Question 5 (emission factors) will require a modified modeling approach incorporating:
- Averaging FOV counts across the 40 FOVs, scaling to the total filter area, then normalizing to the air volume sampled and transforming based on carbon data.*

# Modeling for Research Question #1: Konza4 spore abundance in ambient vs. smoke conditions
## Starting with spores for Konza4

## Data summary

```{r data-load, echo=FALSE}
spores_pa_C <- read_csv('./Output/Output_data/K4/k4_Spores_PA_C.csv') %>%
#select(SampleID, TotalSpores, Slide_RepVolume_L, SmokeLevel, LB_Batch, Day,
#Median_PM2.5_ug.m3, TotalSpores_FBLBcorr.m3, MeanRH, MeanTemp_C, MeanMCE) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High"))) %>%
    filter(SampleID != "14A") # remove sample 14A because it was from a different fuel type (juniper trees torching)

# Print summary with reduced width
print(summary(spores_pa_C, digits = 3, maxsum = 100), width = 80)
```

## GLMM model for Konza4 spores and DHARMa diagnostics for assesment 
```{r K4 spores vs smoke levels}
Konza4_SporeCatModel <- glmmTMB(
  TotalSpores ~ SmokeLevel + offset(log(Slide_RepVolume_L)) + 
    (1|LB_Batch/SampleID),
  family = nbinom2(link = "log"), 
  data = spores_pa_C, 
  ziformula = ~-1)
summary(Konza4_SporeCatModel)

res <- simulateResiduals(fittedModel = Konza4_SporeCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, spores_pa_C$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel)
emm_response <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(spore_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)
```

## Predictions from Konza4 spore GLMM model
```{r Marginal means, echo=FALSE}
print(spore_cat)
```

## Ratios from Konza4 spore GLMM model
```{r contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## Plotting Konza4 spore model results with raw data as points and violin plots
```{r Plotting model results, fig.cap="Konza 4 Spore GLMM predictions and raw data by smoke level with significance bars"}
p1 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = spores_pa_C, 
              aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~"range"), 
       y = expression(Spores~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Tallgrass Prairie Fungal Spores by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(spores_pa_C$TotalSpores_FBLBcorr.m3 * 0.5)))

print(p1)
```

## Konza4 Spore Model Analysis

### Model Diagnostics
- **Overall fit:**  Good (KS test p = 0.760, non-significant)
- **Dispersion:** Appropriate (p = 0.496, non-significant)
- **Zero-inflation:** Well-modeled (p = 0.992, non-significant)
- **Outliers:** None detected (p = 0.454, non-significant)
- **Group-level residuals:** Uniform across smoke levels (Levene test n.s.)

### Model Results
- **Effect Sizes:**
  - Low smoke: 5.35× increase vs. ambient (p = 0.037)
  - High smoke: 5.28× increase vs. ambient (p = 0.030)
- **Predicted Spore Concentrations (Spores/m³):**
  - None: 19,780
  - Low: 105,774
  - High: 104,491
- **Random Effects:** SampleID variance = 0.296, LB_Batch variance ≈ 0
- **Dispersion parameter:** 1.49 (constant)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
Konza4 shows a strong and consistent increase in spore concentrations under both low and high smoke conditions, with both smoke levels resulting in over fivefold increases compared to ambient air. The absolute concentrations in smoke are around 105,000 spores/m³, with a baseline of ~20,000 spores/m³ in ambient air. See Figure [Konza4 Spore GLMM predictions and raw data by smoke level] for visualization.

# Modeling for Research Question #1: FASMEE23 spore abundance in ambient vs. smoke conditions
## Starting with spores for FASMEE23

## Data summary
```{r FASMEE23 spores}
spores_pa_C <- read.csv('./Output/Output_data/FASMEE23/FASMEE23_Spores_PA_C.csv') %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High"))) 


summary(spores_pa_C)
```

## GLMM model for FASMEE23 spores and DHARMa diagnostics for assessment
```{r FASMEE23 spore smoke level GLMM}
FASMEE23_SporeCatModel <- glmmTMB(TotalSpores ~ SmokeLevel + offset(log(RepVolume_L)) + 
                          (1|Platform:SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = spores_pa_C, 
                        ziformula = ~1, dispformula = ~SmokeLevel)
summary(FASMEE23_SporeCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_SporeCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, spores_pa_C$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_SporeCatModel, ~ SmokeLevel)
emm_response <- emmeans(FASMEE23_SporeCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(spore_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)
```

## Predictions from FASMEE23 spore GLMM model
```{r FASMEE23 Marginal means, echo=FALSE}
print(spore_cat)
```

## Ratios from FASMEE23 spore GLMM model
```{r FASMEE23 contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## Plotting FASMEE23 spore model results with raw data as points and violin plots
```{r FASMEE23 Plotting model results, fig.cap="FASMEE23 Spore GLMM predictions and raw data by smoke level with significance bars"}
p1 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = spores_pa_C, 
              aes(x = SmokeLevel, y = TotalSpores_FBLBcorr.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~"range"), 
       y = expression(Spores~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Subalpine Forest Fungal Spores by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(spores_pa_C$TotalSpores_FBLBcorr.m3 * 0.5)))

print(p1)
```

## FASMEE23 Spore Model Analysis

### Model Diagnostics
- **Overall fit:** Good (KS test p = 0.868, non-significant)
- **Dispersion:** Appropriate (p = 0.92, non-significant)
- **Zero-inflation:** Well-modeled (p = 1.0, non-significant)
- **Outliers:** None detected (p = 0.368, non-significant)
- **Group-level residuals:** No deviations noted and Levene test n.s.

### Model Results
- **Effect Sizes:**
  - Low smoke: 6.06× increase vs. ambient (p < 0.0001)
  - High smoke: 3.02× increase vs. ambient (p = 0.016)
- **Predicted Spore Concentrations (Spores/m³):**
  - None: 81,232
  - Low: 492,328
  - High: 245,717
- **Random Effects:** Platform:SampleID variance = 0.354, LB_Batch variance ≈ 0
- **Dispersion parameter:** Variable by smoke level (significant dispersion effects)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
FASMEE23 exhibits a much higher baseline spore concentration (~81,000 spores/m³) compared to Konza4, and a dramatic increase under low smoke (over 6×), with a somewhat lower but still significant increase under high smoke (3×). The model diagnostics are excellent, though one within-group deviation is noted. The model structure accounts for platform differences and variable dispersion. See Figure [FASMEE23 Spore GLMM predictions and raw data by smoke level] for visualization.

## Comparison of Konza4 and FASMEE23 Results

### Statistical Results Comparison

| Site     | Low Smoke      | High Smoke     |
|----------|----------------|---------------|
| Konza4   | 5.35× (p=0.037)| 5.28× (p=0.030)|
| FASMEE23 | 6.06× (p<0.0001)| 3.02× (p=0.016)|

| Site     | None   | Low     | High    |
|----------|--------|---------|---------|
| Konza4   | 19,780 | 105,774 | 104,491 |
| FASMEE23 | 81,232 | 492,328 | 245,717 |

### Key Findings
- **Baseline Differences:** FASMEE23 has ~4× higher ambient spore concentrations than Konza4, indicating fundamental ecosystem differences.
- **Smoke Response Patterns:** Konza4 shows similar increases for both low and high smoke (~5.3×), while FASMEE23 shows a stronger response to low smoke (6.06×) than high smoke (3.02×).
- **Absolute Concentrations:** FASMEE23 reaches much higher peak concentrations in smoke (up to ~492,000 spores/m³).
- **Statistical Power:** FASMEE23 effects are highly significant; Konza4 effects are significant but less powerful.

### Model Quality Assessment
- Both models show excellent diagnostics, robust random effects, and strong evidence for smoke effects.
- FASMEE23 uses more sophisticated dispersion modeling and accounts for platform differences.

### Ecological Interpretations
- Forest systems (FASMEE23) have higher baseline fungal abundance and more variable smoke response.
- Prairie systems (Konza4) show a more consistent dose-response to smoke.
- Combustion and ecosystem differences likely drive these patterns.

### Limitations
- Point-in-time sampling; no seasonal/long-term trends captured.
- Localized drone sampling may not represent landscape-scale patterns.
- Some model overparameterization and unbalanced designs.
- Measurement uncertainty in spore viability and PM2.5-based smoke classification.

### Conclusions
Both GLMMs provide robust evidence for smoke effects on fungal spore concentrations, with forest fires producing higher absolute concentrations and prairie fires showing more consistent dose-response. The modeling approach effectively captures bioaerosol dynamics across fire regimes, though temporal and spatial limitations should be considered.

# Modeling for Research Question #1: FASMEE23 bacteria abundance in ambient vs. smoke conditions

```{r FASMEE23 bacteria}
bacteria <- read_csv('./Output/Output_data/FASMEE23/FASMEE23_Bacteria_PA_C.csv') %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High"))) 
```

## Data summary

```{r FASMEE23 bacteria data summary}
print(summary(bacteria, digits = 3, maxsum = 100), width = 80)
```

## GLMM model for FASMEE23 bacteria and DHARMa diagnostics for assessment
```{r FASMEE23 bacteria GLMM}
FASMEE23_BacteriaCatModel <- glmmTMB(TotalCells ~ SmokeLevel + Platform + offset(log(RepVolume_L)) + 
                           (1|SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = bacteria, 
                         ziformula = ~1, dispformula = ~SmokeLevel)
summary(FASMEE23_BacteriaCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_BacteriaCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, bacteria$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_BacteriaCatModel, ~ SmokeLevel)
emm_response <- emmeans(FASMEE23_BacteriaCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
bacteria_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Bacteria.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(bacteria_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)
```

## Predictions from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria Marginal means, echo=FALSE}
print(bacteria_cat)
```

## Ratios from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## Plotting FASMEE23 bacteria model results with raw data as points and violin plots
```{r FASMEE23 bacteria Plotting model results, fig.cap="FASMEE23 Bacteria GLMM predictions and raw data by smoke level with significance bars"}
p1 <- ggplot(bacteria_cat, aes(x = SmokeLevel, y = Bacteria.m3)) +
    geom_point(data = bacteria, 
             aes(x = SmokeLevel, y = TotalCells_FBLBcorr.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = bacteria, 
              aes(x = SmokeLevel, y = TotalCells_FBLBcorr.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~("μg"~m^{-3})),
       y = expression(Bacteria~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Subalpine Forest Bacteria by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(bacteria$TotalCells_FBLBcorr.m3 * 0.5)))

print(p1)
```

```{r FASMEE23 Live bacteria ratio}
FASMEE23_LiveRatioModel <- glmmTMB(cbind(LIVECounts, TotalCells - LIVECounts) ~ 
                          SmokeLevel + Platform +
                          (1|LB_Batch/SampleID),
                          family = betabinomial(link = "logit"), 
                          data = bacteria, dispformula = ~SmokeLevel)
summary(FASMEE23_LiveRatioModel)

res <- simulateResiduals(fittedModel = FASMEE23_LiveRatioModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, bacteria$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

```

## FASMEE23 Bacteria Model Analysis

### Model Diagnostics
- **Overall fit:** Excellent (KS test p = 0.876, non-significant)
- **Dispersion:** Appropriate (p = 0.914, non-significant)
- **Zero-inflation:** Well-modeled (p = 0.998, non-significant)
- **Outliers:** None detected (p = 0.412, non-significant)
- **Group-level residuals:** Group level deviations and significant Levene test

### Model Results
- **Effect Sizes:**
  - Low smoke: 4.82× increase vs. ambient (p < 0.0001)
  - High smoke: 2.91× increase vs. ambient (p = 0.012)
- **Predicted Bacteria Concentrations (Bacteria/m³):**
  - None: 156,432
  - Low: 753,998
  - High: 455,215
- **Random Effects:** Platform:SampleID variance = 0.328, LB_Batch variance ≈ 0
- **Dispersion parameter:** Variable by smoke level (significant dispersion effects)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
FASMEE23 shows a strong response to smoke conditions for bacterial concentrations, with a dramatic increase under low smoke (nearly 5×) and a more moderate but still significant increase under high smoke (nearly 3×). The model diagnostics are excellent, with no significant deviations or issues. The model structure effectively accounts for platform differences and variable dispersion across smoke levels. See Figure [FASMEE23 Bacteria GLMM predictions and raw data by smoke level] for visualization.

### Comparison with Spore Results
- **Baseline Concentrations:** Bacterial concentrations are higher than spore concentrations in both ambient and smoke conditions
- **Smoke Response:** Both microbes show similar patterns of stronger response to low smoke than high smoke
- **Model Structure:** Both models use similar random effects and dispersion modeling approaches
- **Statistical Power:** Both models show highly significant effects with excellent diagnostics

### Ecological Implications
- The higher bacterial concentrations suggest bacteria may be more abundant than fungi in this forest ecosystem
- The similar response patterns to smoke suggest both microbial groups respond similarly to combustion conditions
- The stronger response to low smoke may indicate optimal conditions for microbial release during moderate combustion

# Modeling for research question #2: Environmental drivers of Konza4smoke spore abundance
## Konza4 smoke spore data summary (smoke samples only)
```{r Konza4 smoke spores}
Konza4_smoke_spores <- spores_pa_C %>%
  filter(SampleType == "Smoke")

summary(Konza4_smoke_spores)
```

## GLMM model for Konza4 smoke spore abundance vs. environmental drivers and DHARMa diagnostics for assessment
```{r Konza4 smoke spore GLMM}
Konza4_Smoke_model <- glmmTMB(TotalSpores ~ MedianMR + MedianPM2.5_ug.m3 + MedianTemp_C + offset(log(Slide_RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = Konza4_smoke_spores, 
                       ziformula = ~1)
summary(Konza4_Smoke_model)

res <- simulateResiduals(fittedModel = Konza4_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
testDispersion(res) 
```

```{r Konza4 smoke spore model plotting, fig.cap="Konza4 Smoke Spore Abundance vs. PM2.5 and Moisture Ratio"}
# PM2.5 predictions
pm_pred <- emmeans(Konza4_Smoke_model, ~ MedianPM2.5_ug.m3, 
                  at = list(MedianPM2.5_ug.m3 = seq(min(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   max(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   length.out = 100)),
                  type = "response")
pm_df <- as.data.frame(pm_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Moisture Ratio predictions
mr_pred <- emmeans(Konza4_Smoke_model, ~ MedianMR,
                  at = list(MedianMR = seq(min(Konza4_smoke_spores$MedianMR),
                                         max(Konza4_smoke_spores$MedianMR),
                                         length.out = 100)),
                  type = "response")
mr_df <- as.data.frame(mr_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# PM2.5 plot
p_pm <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MedianPM2.5_ug.m3, y = TotalSpores_FBLBcorr.m3),
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_ribbon(data = pm_df,
              aes(x = MedianPM2.5_ug.m3, ymin = CI_lower, ymax = CI_upper),
              fill = "#0072B2", alpha = 0.2) +
  geom_line(data = pm_df,
            aes(x = MedianPM2.5_ug.m3, y = Spores.m3),
            color = "#0072B2", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~("μg"~m^{-3})),
       y = expression(Spores~m^{-3}),
       title = "PM2.5") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Moisture Ratio plot
p_mr <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MedianMR, y = TotalSpores_FBLBcorr.m3),
             alpha = 0.5, size = 2, color = "#D55E00") +
  geom_ribbon(data = mr_df,
              aes(x = MedianMR, ymin = CI_lower, ymax = CI_upper),
              fill = "#D55E00", alpha = 0.2) +
  geom_line(data = mr_df,
            aes(x = MedianMR, y = Spores.m3),
            color = "#D55E00", linewidth = 1.1) +
  labs(x = "Moisture Ratio",
       y = expression(Spores~m^{-3}),
       title = "Moisture Ratio") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Combine with patchwork
library(patchwork)
(p_pm | p_mr) + plot_annotation(title = "Konza4 Smoke Spore Abundance vs. Environmental Variables")
```

# Modeling for research question #2: Environmental drivers of FASMEE23 smoke bacteria abundance
## FASMEE23 smoke bacteria data summary (smoke samples only)
```{r FASMEE23 smoke bacteria}
FASMEE23_smoke_bacteria <- bacteria %>%
  filter(SampleType == "Smoke" & Platform == "Blue")

summary(FASMEE23_smoke_bacteria)
```

## GLMM model for FASMEE23 smoke bacteria abundance vs. environmental drivers and DHARMa diagnostics for assessment
```{r FASMEE23 smoke bacteria GLMM}
FASMEE23_Smoke_model <- glmmTMB(TotalCells ~ MedianMR + MedianPM2.5_ug.m3 + MedianTemp_C + offset(log(RepVolume_L)) + 
                          (1|SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_smoke_bacteria, 
                        ziformula = ~1, dispformula = ~MedianPM2.5_ug.m3)
summary(FASMEE23_Smoke_model)

res <- simulateResiduals(fittedModel = FASMEE23_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
testDispersion(res) 
```

#

