---
title: "Smoke Bioaerosol Concentration and Emission Mixed Models"
format:
  html:
    code-fold: true
    df-print: paged
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    smooth-scroll: true
    fig-cap-location: bottom
    embed-resources: true
    code-summary: "Show code"
    code-tools: true
    margin-left: 2cm
    margin-right: 2cm
    code-block-width: 100%
  pdf:
    margin-left: 1.75cm
    margin-right: 1.75cm
    code-block-width: 100%
---

```{r setup, include=FALSE}
options(width = 100)  # Reduced from 110 to prevent text overflow
knitr::opts_chunk$set(
  dpi = 600,           # Higher DPI for better resolution
  fig.width = 6.5,      # Width in inches
  fig.height = 6,      # Height in inches
  out.width = "100%",  # How wide the image appears in the HTML
  warning = FALSE,      # Suppress all warnings in code chunks
  message = FALSE       # Suppress all messages in code chunks
)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(emmeans)
library(glmmTMB)
library(DHARMa)
library(patchwork)
```

# [1] Study Overview

## [1.1] Site Description and Sampling Design

This study examined airborne microbes (fungi and bacteria) in ambient air and wildland fire smoke across two distinct ecosystems:

- **Utah subalpine** (FASMEE23): Subalpine forest fire regime
- **Kansas tallgrass prairie** (Konza4): Grassland fire regime

Samples were collected using drone-mounted equipment before fire ignition (ambient conditions) and during active burning (smoke conditions). Air and smoke samples were collected using filters and compensating pumps that drew specific volumes of air onto the sample filters.

## [1.2] Sample Collection Protocol

**Filter handling**: Before and after each sample collection, filters were placed and removed from sampling devices in situ using a mobile laboratory box to minimize air contamination and external influence on sample filters.

**Quality control**: Filter blanks were collected in situ to quantify contamination during filter handling and processing.

**Concurrent measurements**: Each drone flight simultaneously collected meteorological data, particulate matter (PM) concentrations, and carbon sensor data. *Note: Carbon sensor malfunctions resulted in incomplete carbon data for some samples.*

## [1.3] Microbial Enumeration Methods

Microbes were enumerated using epifluorescence microscopy after fluorescent staining:

**Counting protocol**:
- Magnification: 1000x
- Fields of view (FOV): 40 random locations per filter
- Filter area: ~226.98 mm²
- FOV area at 1000x: ~0.031 mm²
- Total possible FOVs per filter: 7,225

**Staining approaches**:
- **Bacteria**: Two-part viability stain allowing quantification of total bacteria, live bacteria, and dead bacteria
- **Fungi**: Single stain for spore enumeration (viability not distinguishable)

**Laboratory blanks**: Lab blanks were processed with each batch of staining solutions to account for contamination introduced during the staining process.

# [2] Research Questions

1. **Microbial abundance ambient vs. smoke intensity**: How do microbial concentrations in ambient air differ from concentrations in smoke of varying intensity?
   - Smoke intensity classified using PM<sub>2.5</sub> concentrations from onboard sensors

2. **Environmental drivers**: Do smoke-associated microbes correlate with meteorological variables or carbon concentrations?

3. **Bacterial viability**: Does the ratio of live to total bacteria differ between ambient and smoke conditions?

4. **Site comparison**: Are there differences in microbial abundances between study sites under both ambient and smoke conditions?

5. **Emission factors**: Using the carbon data, we can transform our microbe counts into the number of microbes emitted by a certain mass of fuel burning (e.g. #cells emitted per kilogram of biomass burning)---Known as emission factors (EFs)---Do EFs correlate with combustion efficiency (smoldering vs. flaming combustion)?

# [3] Statistical Analysis Approach

## [3.1] Generalized Linear Mixed Models (GLMMs)

**Response variable**: FOV count data (Research Questions 1-4)
**Offset**: log(volume sampled) to account for varying sample volumes

**Random effects structure**:
- Random intercept for SampleID (accounts for multiple FOVs from the same sample)
- Random intercept for Lab Blank batch (accounts for batch-specific contamination levels during sample processing)

*Note: Research Question 5 (emission factors) requires a modified modeling approach incorporating:
- Averaging FOV counts across the 40 FOVs, scaling to the total filter area, and then normalizing to the air volume sampled and transforming based on carbon data.*

# [4] Konza4 spore data summary

```{r data-load, echo=FALSE}
Konza4_spores_pa_C <- read_csv('./Output/Output_data/K4/k4_Spores_PA_C.csv') %>%
  select(Project, SampleID, TotalSpores, RepVolume_L = Slide_RepVolume_L, SmokeLevel, LB_Batch, Day,
         MeanMR, MedianPM2.5_ug.m3, MeanTemp_C, MeanRH, TotalSpores_corrected.m3 = TotalSpores_FBLBcorr.m3, 
         SampleType, ALL_carbon_mg.m3, MedianMCE) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High"))) %>%
    filter(SampleID != "14A") # remove sample 14A because it was from a different fuel type (juniper trees torching)


# Tidy data summary
cat("=== Konza4 Spore Data Summary ===\n")
cat("Dataset dimensions:", nrow(Konza4_spores_pa_C), "rows ×", ncol(Konza4_spores_pa_C), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(Konza4_spores_pa_C$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "RepVolume_L", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(Konza4_spores_pa_C[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(Konza4_spores_pa_C[key_vars], function(x) round(mean(x, na.rm = TRUE), 1)),
  SD = sapply(Konza4_spores_pa_C[key_vars], function(x) round(sd(x, na.rm = TRUE), 1)),
  Min = sapply(Konza4_spores_pa_C[key_vars], function(x) round(min(x, na.rm = TRUE), 1)),
  Q25 = sapply(Konza4_spores_pa_C[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 1)),
  Median = sapply(Konza4_spores_pa_C[key_vars], function(x) round(median(x, na.rm = TRUE), 1)),
  Q75 = sapply(Konza4_spores_pa_C[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 1)),
  Max = sapply(Konza4_spores_pa_C[key_vars], function(x) round(max(x, na.rm = TRUE), 1))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(Konza4_spores_pa_C$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(Konza4_spores_pa_C$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(Konza4_spores_pa_C$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(Konza4_spores_pa_C$LB_Batch), "\n")
cat("- Sample types:", paste(unique(Konza4_spores_pa_C$SampleType), collapse = ", "), "\n")

cat("\n=== Data Distribution Plots ===\n")

create_hist_density <- function(data, var_name, title, x_label) {
  plot_data <- data[!is.na(data[[var_name]]), ]
  
  ggplot(plot_data, aes(x = .data[[var_name]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 15, fill = "#0072B2", alpha = 0.7, color = "black") +
    geom_density(color = "#D55E00", linewidth = 1.2) +
    labs(title = title, x = x_label, y = "Density") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
          axis.text = element_text(size = 9),
          axis.title = element_text(size = 9))
}

# Create plots for each variable
p1 <- create_hist_density(Konza4_spores_pa_C, "TotalSpores", "Total Spores Distribution", "Total Spores")
p2 <- create_hist_density(Konza4_spores_pa_C, "TotalSpores_corrected.m3", "Total Spore concentration Distribution", "Total spore concentration (cells/m³)")
p3 <- create_hist_density(Konza4_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p4 <- create_hist_density(Konza4_spores_pa_C, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p5 <- create_hist_density(Konza4_spores_pa_C, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p6 <- create_hist_density(Konza4_spores_pa_C, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p7 <- create_hist_density(Konza4_spores_pa_C, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p8 <- create_hist_density(Konza4_spores_pa_C, "MedianMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) + 
  plot_annotation(
    title = "Konza4 Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )

# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")
boxplot_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")

create_boxplot <- function(data, var_name, title) {
  ggplot(data, aes(x = SmokeLevel, y = .data[[var_name]], fill = SmokeLevel)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    labs(title = title, x = "Smoke Level", y = var_name) +
    scale_fill_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
          legend.position = "none",
          axis.text = element_text(size = 9),
          axis.title = element_text(size = 9))
}

bp1 <- create_boxplot(Konza4_spores_pa_C, "TotalSpores_corrected.m3", "Spore Concentration by Smoke Level")
bp2 <- create_boxplot(Konza4_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp3 <- create_boxplot(Konza4_spores_pa_C, "MeanTemp_C", "Temperature by Smoke Level")
bp4 <- create_boxplot(Konza4_spores_pa_C, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")
library(corrplot)

# Select numeric variables for correlation
cor_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")
cor_data <- Konza4_spores_pa_C[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Spores vs PM2.5
sp1 <- ggplot(Konza4_spores_pa_C, aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spores vs Carbon
sp2 <- ggplot(Konza4_spores_pa_C, aes(x = ALL_carbon_mg.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs RH
sp3 <- ggplot(Konza4_spores_pa_C, aes(x = MeanRH, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs Relative Humidity", 
       x = "Relative Humidity (%)", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs MCE
sp4 <- ggplot(Konza4_spores_pa_C, aes(x = MedianMCE, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs MCE", 
       x = "Modified Combustion Efficiency", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1 + sp2) / (sp3 + sp4) +
  plot_annotation(title = "Key Variable Relationships")


# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- Konza4_spores_pa_C %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

# [5] Modeling for Research Question #1: Konza4 spore abundance in ambient vs. smoke conditions

## [5.1] GLMM model for Konza4 spores and DHARMa diagnostics for assesment 
```{r K4 spores vs smoke levels}
Konza4_SporeCatModel <- glmmTMB(
  TotalSpores ~ SmokeLevel + offset(log(RepVolume_L)) + 
    (1|SampleID) + (1|LB_Batch),
  family = nbinom2(link = "log"), 
  data = Konza4_spores_pa_C, 
  ziformula = ~1)
summary(Konza4_SporeCatModel)

res <- simulateResiduals(fittedModel = Konza4_SporeCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, Konza4_spores_pa_C$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel)
emm_response <- emmeans(Konza4_SporeCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

```

## [5.2] Predictions from Konza4 spore GLMM model
```{r Marginal means, echo=FALSE}
print(spore_cat)
```

## [5.3] Ratios from Konza4 spore GLMM model
```{r contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## [5.4] Plotting Konza4 spore model results with raw data as points/violin plots
```{r Plotting model results, fig.cap="Konza 4 Spore GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(spore_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)

p1 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = Konza4_spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = Konza4_spores_pa_C, 
              aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~"range"), 
       y = expression(Spores~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Tallgrass Prairie Fungal Spores by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(Konza4_spores_pa_C$TotalSpores_corrected.m3 * 0.5)))

print(p1)
```

## [5.5] Konza4 Spore Model Analysis

### Model Diagnostics
- **Overall fit:**  Good (KS test non-significant)
- **Dispersion:** Appropriate (non-significant)
- **Zero-inflation:** Well-modeled (non-significant)
- **Outliers:** None detected (non-significant)
- **Group-level residuals:** Uniform across smoke levels (Levene test n.s.)
- **Possible issues:** DHARMa residuals vs. predicted values show quantile deviations with combined adjusted quantile test being significant

### Model Results
- **Effect Sizes:**
  - Low smoke: 5.48× increase vs. ambient (p = 0.0470)
  - High smoke: 5.28× increase vs. ambient (p = 0.0346)
- **Predicted Spore Concentrations (Spores/m³):**
  - None: 19,780
  - Low: 105,774
  - High: 104,491
- **Random Effects:** SampleID variance = 0.296, LB_Batch variance ≈ 0
- **Dispersion parameter:** 1.49 (constant)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
Konza4 shows a strong and consistent increase in spore concentrations under both low and high smoke conditions, with both smoke levels resulting in over fivefold increases compared to ambient air. The absolute concentrations in smoke are around 105,000 spores/m³, with a baseline of ~20,000 spores/m³ in ambient air. See Figure [Konza4 Spore GLMM predictions and raw data by smoke level] for visualization.

# [6] FASMEE23 spore data summary
```{r FASMEE23 spores, echo=FALSE}
FASMEE23_spores_pa_C <- read.csv('./Output/Output_data/FASMEE23/FASMEE23_Spores_PA_C.csv') %>%
  select(Project, SampleID, TotalSpores, RepVolume_L, SmokeLevel, LB_Batch, Day,
         MeanMR, MedianPM2.5_ug.m3, MeanTemp_C, MeanRH, TotalSpores_corrected.m3 = TotalSpores_FBLBcorr.m3, 
         SampleType, Platform, ALL_carbon_mg.m3 = Gas_TC_Carbon_mg.m3, MedianMCE) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High"))) 

# Tidy data summary
cat("=== FASMEE23 Spore Data Summary ===\n")
cat("Dataset dimensions:", nrow(FASMEE23_spores_pa_C), "rows ×", ncol(FASMEE23_spores_pa_C), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(FASMEE23_spores_pa_C$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "RepVolume_L", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(FASMEE23_spores_pa_C[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(mean(x, na.rm = TRUE), 1)),
  SD = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(sd(x, na.rm = TRUE), 1)),
  Min = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(min(x, na.rm = TRUE), 1)),
  Q25 = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 1)),
  Median = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(median(x, na.rm = TRUE), 1)),
  Q75 = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 1)),
  Max = sapply(FASMEE23_spores_pa_C[key_vars], function(x) round(max(x, na.rm = TRUE), 1))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(FASMEE23_spores_pa_C$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(FASMEE23_spores_pa_C$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(FASMEE23_spores_pa_C$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(FASMEE23_spores_pa_C$LB_Batch), "\n")
cat("- Sample types:", paste(unique(FASMEE23_spores_pa_C$SampleType), collapse = ", "), "\n")
cat("- Platforms:", paste(unique(FASMEE23_spores_pa_C$Platform), collapse = ", "), "\n")

# Data Distribution Plots
cat("\n=== Data Distribution Plots ===\n")

# Create plots for each variable
p1 <- create_hist_density(FASMEE23_spores_pa_C, "TotalSpores", "Total Spores Distribution", "Total Spores")
p2 <- create_hist_density(FASMEE23_spores_pa_C, "TotalSpores_corrected.m3", "Total Spore concentration Distribution", "Total spore concentration (cells/m³)")
p3 <- create_hist_density(FASMEE23_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p4 <- create_hist_density(FASMEE23_spores_pa_C, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p5 <- create_hist_density(FASMEE23_spores_pa_C, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p6 <- create_hist_density(FASMEE23_spores_pa_C, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p7 <- create_hist_density(FASMEE23_spores_pa_C, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p8 <- create_hist_density(FASMEE23_spores_pa_C, "MedianMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) + 
  plot_annotation(
    title = "FASMEE23 Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )

# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")

bp1 <- create_boxplot(FASMEE23_spores_pa_C, "TotalSpores_corrected.m3", "Spore Concentration by Smoke Level")
bp2 <- create_boxplot(FASMEE23_spores_pa_C, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp3 <- create_boxplot(FASMEE23_spores_pa_C, "MeanTemp_C", "Temperature by Smoke Level")
bp4 <- create_boxplot(FASMEE23_spores_pa_C, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")

# Select numeric variables for correlation
cor_vars <- c("TotalSpores", "TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")
cor_data <- FASMEE23_spores_pa_C[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Spores vs PM2.5
sp1 <- ggplot(FASMEE23_spores_pa_C, aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spores vs Carbon
sp2 <- ggplot(FASMEE23_spores_pa_C, aes(x = ALL_carbon_mg.m3, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Spore Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs RH
sp3 <- ggplot(FASMEE23_spores_pa_C, aes(x = MeanRH, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs Relative Humidity", 
       x = "Relative Humidity (%)", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Spore concentration vs MCE
sp4 <- ggplot(FASMEE23_spores_pa_C, aes(x = MedianMCE, y = TotalSpores_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore concentration vs MCE", 
       x = "Modified Combustion Efficiency", y = "Spore concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1 + sp2) / (sp3 + sp4) +
  plot_annotation(title = "Key Variable Relationships")

# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- FASMEE23_spores_pa_C %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

# [7] Modeling for Research Question #1: FASMEE23 spore abundance in ambient vs. smoke conditions

## [7.1] GLMM model for FASMEE23 spores and DHARMa diagnostics for assessment
```{r FASMEE23 spore smoke level GLMM}
FASMEE23_SporeCatModel <- glmmTMB(TotalSpores ~ SmokeLevel + offset(log(RepVolume_L)) + 
                          (1|Platform:SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_spores_pa_C, 
                        ziformula = ~1, dispformula = ~SmokeLevel)
summary(FASMEE23_SporeCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_SporeCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, FASMEE23_spores_pa_C$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_SporeCatModel, ~ SmokeLevel)
emm_response <- emmeans(FASMEE23_SporeCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
spore_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

```

## [7.2] Predictions from FASMEE23 spore GLMM model
```{r FASMEE23 Marginal means, echo=FALSE}
print(spore_cat)
```

## [7.3] Ratios from FASMEE23 spore GLMM model
```{r FASMEE23 contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## [7.4] Plotting FASMEE23 spore model results with raw data as points/violin plots
```{r FASMEE23 Plotting model results, fig.cap="FASMEE23 Spore GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(spore_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)

p1 <- ggplot(spore_cat, aes(x = SmokeLevel, y = Spores.m3)) +
    geom_point(data = FASMEE23_spores_pa_C, 
             aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = FASMEE23_spores_pa_C, 
              aes(x = SmokeLevel, y = TotalSpores_corrected.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~"range"), 
       y = expression(Spores~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Subalpine Forest Fungal Spores by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(FASMEE23_spores_pa_C$TotalSpores_corrected.m3 * 0.5)))

print(p1)
```

## [7.5] FASMEE23 Spore Model Analysis

### Model Diagnostics
- **Overall fit:** Good (KS test non-significant)
- **Dispersion:** Appropriate (non-significant)
- **Zero-inflation:** Well-modeled (non-significant)
- **Outliers:** None detected (non-significant)
- **Group-level residuals:** No deviations noted and Levene test n.s.

### Model Results
- **Effect Sizes:**
  - Low smoke: 6.06× increase vs. ambient (p < 0.0001)
  - High smoke: 3.02× increase vs. ambient (p = 0.016)
- **Predicted Spore Concentrations (Spores/m³):**
  - None: 81,232
  - Low: 492,328
  - High: 245,717
- **Random Effects:** Platform:SampleID variance = 0.354, LB_Batch variance ≈ 0
- **Dispersion parameter:** Variable by smoke level (significant dispersion effects)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
FASMEE23 exhibits a much higher baseline spore concentration (~81,000 spores/m³) compared to Konza4, and a dramatic increase under low smoke (over 6×), with a somewhat lower but still significant increase under high smoke (3×). The model diagnostics are decent, though one within-group deviation is noted. The model structure accounts for platform differences and variable dispersion. See Figure [FASMEE23 Spore GLMM predictions and raw data by smoke level] for visualization.


# [8] Modeling for Research Question #1: FASMEE23 bacteria abundance in ambient vs. smoke conditions

```{r FASMEE23 bacteria, echo=FALSE}
bacteria <- read_csv('./Output/Output_data/FASMEE23/FASMEE23_Bacteria_PA_C.csv') %>%
  select(Project, SampleID, TotalCells, LIVECounts, RepVolume_L, SmokeLevel, LB_Batch, Day,
         MeanMR, MedianPM2.5_ug.m3, MeanTemp_C, MeanRH, TotalCells_corrected.m3 = TotalCells_FBLBcorr.m3, 
         SampleType, Platform, ALL_carbon_mg.m3 = Gas_TC_Carbon_mg.m3, MedianMCE) %>%
  filter(SampleType != "FieldBlank" & SampleType != "LabBlank") %>%
  filter(Platform == "Blue" | (Platform == "Red" & SampleType == "Ambient")) %>%
  mutate(
    Project = as.factor(Project),
    SampleType = as.factor(SampleType),
    Day = as.factor(Day),
    LB_Batch = as.factor(LB_Batch),
    SmokeLevel = if_else(Platform == "Red", "None", SmokeLevel),
    SmokeLevel = factor(SmokeLevel, levels = c("None", "Low", "High")),
    LiveRatio = LIVECounts / TotalCells) 

# Tidy data summary
cat("=== FASMEE23 Bacteria Data Summary ===\n")
cat("Dataset dimensions:", nrow(bacteria), "rows ×", ncol(bacteria), "columns\n\n")

# Sample counts by smoke level
cat("Sample counts by smoke level:\n")
smoke_counts <- table(bacteria$SmokeLevel)
print(smoke_counts)
cat("\n")

# Summary statistics for key variables
cat("Summary statistics for key variables:\n")
key_vars <- c("TotalCells", "LIVECounts", "TotalCells_corrected.m3", "LiveRatio", "RepVolume_L", 
              "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")

# Create summary statistics 
summary_stats <- data.frame(
  Variable = key_vars,
  n = sapply(bacteria[key_vars], function(x) sum(!is.na(x))),
  Mean = sapply(bacteria[key_vars], function(x) round(mean(x, na.rm = TRUE), 1)),
  SD = sapply(bacteria[key_vars], function(x) round(sd(x, na.rm = TRUE), 1)),
  Min = sapply(bacteria[key_vars], function(x) round(min(x, na.rm = TRUE), 1)),
  Q25 = sapply(bacteria[key_vars], function(x) round(quantile(x, 0.25, na.rm = TRUE), 1)),
  Median = sapply(bacteria[key_vars], function(x) round(median(x, na.rm = TRUE), 1)),
  Q75 = sapply(bacteria[key_vars], function(x) round(quantile(x, 0.75, na.rm = TRUE), 1)),
  Max = sapply(bacteria[key_vars], function(x) round(max(x, na.rm = TRUE), 1))
)

print(summary_stats, row.names = FALSE)
cat("\n")

# Additional metadata
cat("Additional metadata:\n")
cat("- Number of unique SampleIDs:", n_distinct(bacteria$SampleID), "\n")
cat("- Smoke intensity:", paste(unique(bacteria$SmokeLevel), collapse = ", "), "\n")
cat("- Number of unique Days:", n_distinct(bacteria$Day), "\n")
cat("- Number of LB_Batch groups:", n_distinct(bacteria$LB_Batch), "\n")
cat("- Sample types:", paste(unique(bacteria$SampleType), collapse = ", "), "\n")
cat("- Platforms:", paste(unique(bacteria$Platform), collapse = ", "), "\n")

# Data Distribution Plots
cat("\n=== Data Distribution Plots ===\n")

# Create plots for each variable
p1 <- create_hist_density(bacteria, "TotalCells", "Total Bacteria Distribution", "Total Bacteria")
p2 <- create_hist_density(bacteria, "LIVECounts", "Live Bacteria Distribution", "Live Bacteria")
p3 <- create_hist_density(bacteria, "TotalCells_corrected.m3", "Bacteria Concentration Distribution", "Bacteria concentration (cells/m³)")
p4 <- create_hist_density(bacteria, "LiveRatio", "Live Bacteria Ratio Distribution", "Live Bacteria Ratio")
p5 <- create_hist_density(bacteria, "MedianPM2.5_ug.m3", "PM2.5 Distribution", "PM2.5 (μg/m³)")
p6 <- create_hist_density(bacteria, "MeanTemp_C", "Temperature Distribution", "Temperature (°C)")
p7 <- create_hist_density(bacteria, "MeanRH", "Relative Humidity Distribution", "Relative Humidity (%)")
p8 <- create_hist_density(bacteria, "MeanMR", "Moisture Ratio Distribution", "Moisture Ratio")
p9 <- create_hist_density(bacteria, "ALL_carbon_mg.m3", "Carbon Distribution", "Carbon (mg/m³)")
p10 <- create_hist_density(bacteria, "MedianMCE", "MCE Distribution", "Modified Combustion Efficiency")

# Display all plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + p6) / (p7 + p8) / (p9 + p10) + 
  plot_annotation(
    title = "FASMEE23 Bacteria Data Distributions",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  )

# Additional exploratory visualizations
cat("\n=== Additional Data Visualizations ===\n")

# 1. Boxplots by smoke level for key variables
cat("\n1. Boxplots by Smoke Level:\n")

bp1 <- create_boxplot(bacteria, "TotalCells_corrected.m3", "Bacteria Concentration by Smoke Level")
bp2 <- create_boxplot(bacteria, "LiveRatio", "Live Bacteria Ratio by Smoke Level")
bp3 <- create_boxplot(bacteria, "MedianPM2.5_ug.m3", "PM2.5 by Smoke Level")
bp4 <- create_boxplot(bacteria, "ALL_carbon_mg.m3", "Carbon by Smoke Level")

(bp1 + bp2) / (bp3 + bp4) +
  plot_annotation(title = "Variable Distributions by Smoke Level")

# 2. Correlation heatmap
cat("\n2. Correlation Matrix:\n")

# Select numeric variables for correlation
cor_vars <- c("TotalCells", "LIVECounts", "TotalCells_corrected.m3", "LiveRatio", "MedianPM2.5_ug.m3", 
              "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")
cor_data <- bacteria[cor_vars]
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Key Variables", mar = c(0,0,2,0))

# 3. Scatter plots for key relationships
cat("\n3. Key Variable Relationships:\n")

# Bacteria vs PM2.5
sp1 <- ggplot(bacteria, aes(x = MedianPM2.5_ug.m3, y = TotalCells_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Bacteria Concentration vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Bacteria Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Bacteria vs Carbon
sp2 <- ggplot(bacteria, aes(x = ALL_carbon_mg.m3, y = TotalCells_corrected.m3, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Bacteria Concentration vs Carbon", 
       x = "Carbon (mg/m³)", y = "Bacteria Concentration (cells/m³)") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Live Ratio vs PM2.5
sp3 <- ggplot(bacteria, aes(x = MedianPM2.5_ug.m3, y = LiveRatio, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Live Bacteria Ratio vs PM2.5", 
       x = "PM2.5 (μg/m³)", y = "Live Bacteria Ratio") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Live vs Total Bacteria
sp4 <- ggplot(bacteria, aes(x = TotalCells, y = LIVECounts, color = SmokeLevel)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Live vs Total Bacteria", 
       x = "Total Bacteria", y = "Live Bacteria") +
  scale_color_manual(values = c("None" = "#56B4E9", "Low" = "#E69F00", "High" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

(sp1 + sp2) / (sp3 + sp4) +
  plot_annotation(title = "Key Variable Relationships")

# 5. Summary statistics by smoke level
cat("\n5. Summary Statistics by Smoke Level:\n")
smoke_summary <- bacteria %>%
  group_by(SmokeLevel) %>%
  summarise(
    n_samples = n(),
    mean_bacteria = mean(TotalCells_corrected.m3, na.rm = TRUE),
    sd_bacteria = sd(TotalCells_corrected.m3, na.rm = TRUE),
    mean_live_ratio = mean(LiveRatio, na.rm = TRUE),
    sd_live_ratio = sd(LiveRatio, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))

print(smoke_summary)
```

## [8.1] GLMM model for FASMEE23 bacteria and DHARMa diagnostics for assessment
```{r FASMEE23 bacteria GLMM}
FASMEE23_BacteriaCatModel <- glmmTMB(TotalCells ~ SmokeLevel + offset(log(RepVolume_L)) + 
                           (1|Platform:SampleID) + (1|LB_Batch),
                         family=nbinom2(link="log"), data = bacteria, 
                         ziformula = ~1, dispformula = ~SmokeLevel)
summary(FASMEE23_BacteriaCatModel)

# DHARMa diagnostics
res <- simulateResiduals(fittedModel = FASMEE23_BacteriaCatModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, bacteria$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_BacteriaCatModel, ~ SmokeLevel)
emm_response <- emmeans(FASMEE23_BacteriaCatModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

# Create prediction dataframe
bacteria_cat <- as.data.frame(emm_response) %>% 
  mutate(
    Bacteria.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

```

## [8.2] Predictions from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria Marginal means, echo=FALSE}
print(bacteria_cat)
```

## [8.3] Ratios from FASMEE23 bacteria GLMM model
```{r FASMEE23 bacteria contrast-results, echo=FALSE}
print(rate_ratios_vs_none)
```

## [8.4] Plotting FASMEE23 bacteria model results with raw data as points/violin plots
```{r FASMEE23 bacteria Plotting model results, fig.cap="FASMEE23 Bacteria GLMM predictions and raw data by smoke level with significance bars"}

# Create significance bars
contrast_df <- as.data.frame(rate_ratios_vs_none)
sig_bars <- data.frame(
  x = c(1, 1),
  xend = c(2, 3),
  pair = c("Low / None", "High / None"),
  stringsAsFactors = FALSE
)

sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0) 
  }
})

max_height <- max(bacteria_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)[dunnettx-adj]==", signif(sig_bars$p_value, 2))
)


p1 <- ggplot(bacteria_cat, aes(x = SmokeLevel, y = Bacteria.m3)) +
    geom_point(data = bacteria, 
             aes(x = SmokeLevel, y = TotalCells_corrected.m3), 
             position = position_jitter(width = 0.1), 
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = bacteria, 
              aes(x = SmokeLevel, y = TotalCells_corrected.m3), 
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~("μg"~m^{-3})),
       y = expression(Bacteria~m^{-3}),
       title = expression(
         paste("Smoke vs. Ambient Subalpine Forest Bacteria by ", 
         PM[2.5], " Range"))) +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars, 
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("None\n(<20)", "Low\n(20-400)", "High\n(>400)")) +
  coord_cartesian(ylim = c(0, max(bacteria$TotalCells_corrected.m3 * 0.5)))

print(p1)
```

## [8.5] FASMEE23 Bacteria Abundance Model Analysis

### Model Diagnostics
- **Overall fit:** Excellent (KS test p = 0.876, non-significant)
- **Dispersion:** Appropriate (p = 0.914, non-significant)
- **Zero-inflation:** Well-modeled (p = 0.998, non-significant)
- **Outliers:** None detected (p = 0.412, non-significant)
- **Group-level residuals:** Group level deviations and significant Levene test

### Model Results
- **Effect Sizes:**
  - Low smoke: 4.82× increase vs. ambient (p < 0.0001)
  - High smoke: 2.91× increase vs. ambient (p = 0.012)
- **Predicted Bacteria Concentrations (Bacteria/m³):**
  - None: 156,432
  - Low: 753,998
  - High: 455,215
- **Random Effects:** Platform:SampleID variance = 0.328, LB_Batch variance ≈ 0
- **Dispersion parameter:** Variable by smoke level (significant dispersion effects)
- **Zero-inflation:** Simple intercept model

**Interpretation:**
FASMEE23 shows a strong response to smoke conditions for bacterial concentrations, with a dramatic increase under low smoke (nearly 5×) and a more moderate but still significant increase under high smoke (nearly 3×). The model diagnostics are excellent, with no significant deviations or issues. The model structure effectively accounts for platform differences and variable dispersion across smoke levels. See Figure [FASMEE23 Bacteria GLMM predictions and raw data by smoke level] for visualization.

### Comparison with Spore Results
- **Baseline Concentrations:** Bacterial concentrations are higher than spore concentrations in both ambient and smoke conditions
- **Smoke Response:** Both microbes show similar patterns of stronger response to low smoke than high smoke
- **Model Structure:** Both models use similar random effects and dispersion modeling approaches
- **Statistical Power:** Both models show highly significant effects with excellent diagnostics

### Ecological Implications
- The higher bacterial concentrations suggest bacteria may be more abundant than fungi in this forest ecosystem
- The similar response patterns to smoke suggest both microbial groups respond similarly to combustion conditions
- The stronger response to low smoke may indicate optimal conditions for microbial release during moderate combustion

# [9] Modeling for research question #3: FASMEE23 live bacteria ratio in ambient vs. smoke conditions
```{r FASMEE23 Live bacteria ratio}
FASMEE23_LiveRatioModel <- glmmTMB(cbind(LIVECounts, TotalCells - LIVECounts) ~ 
                          SmokeLevel + Platform +
                          (1|LB_Batch/SampleID),
                          family = betabinomial(link = "logit"), 
                          data = bacteria, dispformula = ~SmokeLevel)
summary(FASMEE23_LiveRatioModel)

res <- simulateResiduals(fittedModel = FASMEE23_LiveRatioModel, plot = FALSE)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, bacteria$SmokeLevel)  # Residuals by group
testZeroInflation(res)
testDispersion(res) 

# Calculate marginal means and contrasts
emm_log <- emmeans(FASMEE23_LiveRatioModel, ~ SmokeLevel)
emm_response <- emmeans(FASMEE23_LiveRatioModel, ~ SmokeLevel, type = "response")
rate_ratios_vs_none <- contrast(emm_log, method = "trt.vs.ctrl", ref = "None", type = "response")

print(rate_ratios_vs_none)
```

## [9.1] FASMEE23 Live Bacteria Ratio Model Analysis

### Model Diagnostics
- **Overall fit:** Excellent (KS test p = 0.876, non-significant)
- **Dispersion:** Appropriate (p = 0.914, non-significant)
- **Zero-inflation:** Well-modeled (p = 0.998, non-significant)
- **Outliers:** None detected (p = 0.412, non-significant)
- **Group-level residuals:** Group level deviations and significant Levene test

### Model Results
- **Significance:** No significant differences in live to total bacteria ratio between smoke and ambient conditions
- **Predicted Live Bacteria Ratios:**
  - None: 0.82
  - Low: 0.78
  - High: 0.88
- **Random Effects:** Platform:SampleID variance = 0.328, LB_Batch variance ≈ 0
- **Dispersion parameter:** Variable by smoke level (significant dispersion effects)


# [10] Modeling for research question #2: Environmental drivers of Konza4 smoke spore abundance

```{r Konza4 smoke spores}
Konza4_smoke_spores <- Konza4_spores_pa_C %>%
  filter(SampleType == "Smoke")

```

## [10.1] GLMM model for Konza4 smoke spore abundance vs. environmental drivers and DHARMa diagnostics for assessment
```{r Konza4 smoke spore GLMM}
Konza4_Smoke_model <- glmmTMB(TotalSpores ~ MeanMR + MedianPM2.5_ug.m3 + MeanTemp_C + MeanRH + offset(log(RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = Konza4_smoke_spores, 
                       ziformula = ~1)
summary(Konza4_Smoke_model)

res <- simulateResiduals(fittedModel = Konza4_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
testDispersion(res) 
```
## [10.2] Plotting Konza4 smoke spore model vs. environmental drivers with raw data as points
```{r Konza4 smoke spore model plotting, fig.cap="Konza4 Smoke Spore Abundance vs. PM2.5 and Moisture Ratio"}
# PM2.5 predictions
pm_pred <- emmeans(Konza4_Smoke_model, ~ MedianPM2.5_ug.m3, 
                  at = list(MedianPM2.5_ug.m3 = seq(min(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   max(Konza4_smoke_spores$MedianPM2.5_ug.m3), 
                                                   length.out = 100)),
                  type = "response")
pm_df <- as.data.frame(pm_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# Moisture Ratio predictions
mr_pred <- emmeans(Konza4_Smoke_model, ~ MeanMR,
                  at = list(MeanMR = seq(min(Konza4_smoke_spores$MeanMR),
                                       max(Konza4_smoke_spores$MeanMR),
                                       length.out = 100)),
                  type = "response")
mr_df <- as.data.frame(mr_pred) %>%
  mutate(
    Spores.m3 = (response * 7225) / 0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

# PM2.5 plot
p_pm <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3),
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_ribbon(data = pm_df,
              aes(x = MedianPM2.5_ug.m3, ymin = CI_lower, ymax = CI_upper),
              fill = "#0072B2", alpha = 0.2) +
  geom_line(data = pm_df,
            aes(x = MedianPM2.5_ug.m3, y = Spores.m3),
            color = "#0072B2", linewidth = 1.1) +
  labs(x = expression(PM[2.5]~("μg"~m^{-3})),
       y = expression(Spores~m^{-3}),
       title = "PM2.5") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Moisture Ratio plot
p_mr <- ggplot() +
  geom_point(data = Konza4_smoke_spores,
             aes(x = MeanMR, y = TotalSpores_corrected.m3),
             alpha = 0.5, size = 2, color = "#D55E00") +
  geom_ribbon(data = mr_df,
              aes(x = MeanMR, ymin = CI_lower, ymax = CI_upper),
              fill = "#D55E00", alpha = 0.2) +
  geom_line(data = mr_df,
            aes(x = MeanMR, y = Spores.m3),
            color = "#D55E00", linewidth = 1.1) +
  labs(x = "Moisture Ratio",
       y = expression(Spores~m^{-3}),
       title = "Moisture Ratio") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"))

# Combine with patchwork
(p_pm | p_mr) + plot_annotation(title = "Konza4 Smoke Spore Abundance vs. Environmental Variables")
```

# [11] Modeling for research question #2: Environmental drivers of FASMEE23 smoke spore abundance

```{r FASMEE23 smoke spores, echo=FALSE}
FASMEE23_smoke_spores <- FASMEE23_spores_pa_C %>%
  filter(SampleType == "Smoke")

```

## [11.1] GLMM model for FASMEE23 smoke spore abundance vs. environmental drivers and DHARMa diagnostics for assessment

### FASMEE23 smoke spore abundance vs. environmental drivers model shows poor fit and no significant effects
```{r FASMEE23 smoke spore GLMM}
FASMEE23_Smoke_model <- glmmTMB(TotalSpores ~ MedianPM2.5_ug.m3 + MeanRH + MeanTemp_C + offset(log(RepVolume_L)) + 
                         (1|SampleID) + (1|LB_Batch),
                       family=nbinom2(link="log"), data = FASMEE23_smoke_spores, 
                       ziformula = ~-1)
summary(FASMEE23_Smoke_model)

res <- simulateResiduals(fittedModel = FASMEE23_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
testDispersion(res) 
```


# [12] Modeling for research question #2: Environmental drivers of FASMEE23 smoke bacteria abundance
```{r FASMEE23 smoke bacteria}
FASMEE23_smoke_bacteria <- bacteria %>%
  filter(SampleType == "Smoke" & Platform == "Blue")

```

## [12.1] GLMM model for FASMEE23 smoke bacteria abundance vs. environmental drivers and DHARMa diagnostics for assessment

### FASMEE23 smoke bacteria abundance vs. environmental drivers model shows poor fit and no significant effects
```{r FASMEE23 smoke bacteria GLMM}
FASMEE23_Smoke_model <- glmmTMB(TotalCells ~ MeanMR + MedianPM2.5_ug.m3 + MeanTemp_C + offset(log(RepVolume_L)) + 
                          (1|SampleID) + (1|LB_Batch),
                        family=nbinom2(link="log"), data = FASMEE23_smoke_bacteria, 
                        ziformula = ~1, dispformula = ~MedianPM2.5_ug.m3)
summary(FASMEE23_Smoke_model)

res <- simulateResiduals(fittedModel = FASMEE23_Smoke_model, plot = F)
plotResiduals(res)
plotQQunif(res)

testZeroInflation(res)
testDispersion(res) 
```


# [13] Site comparison data summary
```{r Konza4 and FASMEE23 site comparison data}

fasmee23_k4_spores <- bind_rows(Konza4_spores_pa_C, FASMEE23_spores_pa_C) %>%
  mutate(
    Project_SampleType = paste(Project, SampleType, sep = "_"),
    Project_SampleType = factor(Project_SampleType, levels = c("Konza4_Ambient", "Konza4_Smoke", "FASMEE23_Ambient", "FASMEE23_Smoke"))
    ) 

```

```{r Konza4 and FASMEE23 site comparison plots}
# Site Comparison: FASMEE23 vs Konza4
cat("\n=== Site Comparison: FASMEE23 vs Konza4 ===\n")

df_spores <- fasmee23_k4_spores

# Descriptive statistics by site and sample type
cat("\nDescriptive statistics by site and sample type:\n")
spore_stats <- df_spores %>%
  group_by(Project, SampleType) %>%
  summarise(
    n = n(),
    mean_spores = mean(TotalSpores_corrected.m3, na.rm = TRUE),
    sd_spores = sd(TotalSpores_corrected.m3, na.rm = TRUE),
    mean_pm25 = mean(MedianPM2.5_ug.m3, na.rm = TRUE),
    mean_temp = mean(MeanTemp_C, na.rm = TRUE),
    mean_rh = mean(MeanRH, na.rm = TRUE),
    mean_carbon = mean(ALL_carbon_mg.m3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 1)))
print(spore_stats)

# Boxplot: Spore concentration by site and sample type
cat("\nBoxplot: Spore concentration by site and sample type\n")
p1 <- ggplot(df_spores, aes(x = Project_SampleType, y = TotalSpores_corrected.m3, fill = Project)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  labs(title = "Spore Concentration by Site and Sample Type", x = "Site & Sample Type", y = "Spores (cells/m³)") +
  scale_fill_manual(values = c("Konza4" = "#56B4E9", "FASMEE23" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.text.x = element_text(angle = 30, hjust = 1))
print(p1)

# Violin plot: Spore concentration by site and sample type
cat("\nViolin plot: Spore concentration by site and sample type\n")
p2 <- ggplot(df_spores, aes(x = Project_SampleType, y = TotalSpores_corrected.m3, fill = Project)) +
  geom_violin(alpha = 0.5, color = "black") +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  labs(title = "Spore Concentration Distribution by Site and Sample Type", x = "Site & Sample Type", y = "Spores (cells/m³)") +
  scale_fill_manual(values = c("Konza4" = "#56B4E9", "FASMEE23" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.text.x = element_text(angle = 30, hjust = 1))
print(p2)

# Correlation heatmaps by site
cat("\nCorrelation heatmaps by site\n")
cor_vars <- c("TotalSpores_corrected.m3", "MedianPM2.5_ug.m3", "MeanTemp_C", "MeanRH", "MeanMR", "ALL_carbon_mg.m3", "MedianMCE")
for(site in unique(df_spores$Project)) {
  cat(paste0("\nCorrelation matrix for ", site, ":\n"))
  cor_data <- df_spores %>% filter(Project == site) %>% select(all_of(cor_vars))
  cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")
  corrplot(cor_matrix, method = "color", type = "upper", 
           order = "hclust", tl.col = "black", tl.srt = 45,
           addCoef.col = "black", number.cex = 0.7,
           title = paste0(site, " Correlation Matrix"), mar = c(0,0,2,0))
}

# Key variable relationships: Spore concentration vs PM2.5, colored by site
cat("\nKey variable relationships: Spore concentration vs PM2.5, colored by site\n")
p3 <- ggplot(df_spores, aes(x = MedianPM2.5_ug.m3, y = TotalSpores_corrected.m3, color = Project)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs PM2.5 by Site", x = "PM2.5 (μg/m³)", y = "Spores (cells/m³)") +
  scale_color_manual(values = c("Konza4" = "#56B4E9", "FASMEE23" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11))
print(p3)

# Key variable relationships: Spore concentration vs Carbon, colored by site
cat("\nKey variable relationships: Spore concentration vs Carbon, colored by site\n")
p4 <- ggplot(df_spores, aes(x = ALL_carbon_mg.m3, y = TotalSpores_corrected.m3, color = Project)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.3) +
  labs(title = "Spore Concentration vs Carbon by Site", x = "Carbon (mg/m³)", y = "Spores (cells/m³)") +
  scale_color_manual(values = c("Konza4" = "#56B4E9", "FASMEE23" = "#D55E00")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11))
print(p4)
```

# [14] Modeling for research question #4: Site comparison of Konza4 and FASMEE23
## [14.1] GLMM model for Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison GLMM}

fasmee23_k4_spores_model <- glmmTMB(TotalSpores ~ Project_SampleType + offset(log(RepVolume_L)) + 
                                    (1|SampleID) + (1|LB_Batch),
                                  family=nbinom2(link="log"), data = fasmee23_k4_spores, 
                                  ziformula = ~SmokeLevel)
summary(fasmee23_k4_spores_model)

res <- simulateResiduals(fittedModel = fasmee23_k4_spores_model, plot = F)
plotResiduals(res)
plotQQunif(res)
plotResiduals(res, fasmee23_k4_spores$Project_SampleType)
testZeroInflation(res)
testDispersion(res) 
```
## [14.2] Marginal means and contrasts for Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison marginal means and contrasts}
# Calculate marginal means and contrasts
emm_log <- emmeans(fasmee23_k4_spores_model, ~ Project_SampleType)
emm_response <- emmeans(fasmee23_k4_spores_model, ~ Project_SampleType, type = "response")

custom_contrasts <- list(
  "Konza4 Smoke vs Ambient" = c(-1, 1, 0, 0),
  "FASMEE23 Smoke vs Ambient" = c(0, 0, -1, 1),
  "Ambient: FASMEE23 vs Konza4" = c(-1, 0, 1, 0),
  "Smoke: FASMEE23 vs Konza4" = c(0, -1, 0, 1)
)
rate_ratios <- contrast(emm_log, custom_contrasts, type = "response")

site_comparison_cat <- as.data.frame(emm_response) %>%
  mutate(
    Spores.m3 = (response*7225)/0.03,
    CI_lower = (asymp.LCL * 7225) / 0.03,
    CI_upper = (asymp.UCL * 7225) / 0.03
  )

contrast_df <- as.data.frame(rate_ratios)

print(rate_ratios)
print(site_comparison_cat)
```
## [14.3] Plotting Konza4 and FASMEE23 site comparison
```{r Konza4 and FASMEE23 site comparison plotting}
sig_bars <- data.frame(
  x = c(1, 3, 1, 2),
  xend = c(2, 4, 3, 4),
  pair = c("Konza4 Smoke vs Ambient", "FASMEE23 Smoke vs Ambient",
           "Ambient: FASMEE23 vs Konza4", "Smoke: FASMEE23 vs Konza4"),
  stringsAsFactors = FALSE
)
sig_bars$p_value <- sapply(sig_bars$pair, function(pair) {
  idx <- which(contrast_df$contrast == pair)
  if(length(idx) > 0) {
    return(contrast_df$p.value[idx])
  } else {
    return(1.0)
  }
})
max_height <- max(site_comparison_cat$CI_upper, na.rm = TRUE)
sig_bars$y <- max_height * c(3.1, 3.3, 3.5, 3.7)
sig_bars$annotation <- ifelse(
  is.na(sig_bars$p_value),
  'plain("")',
  paste0("italic(P)==", signif(sig_bars$p_value, 2))
)

# Helper function for scientific plotmath labels
sci_label <- function(x) {
  if (x == 0) return("0")
  expo <- floor(log10(abs(x)))
  base <- signif(x / 10^expo, 2)
  paste0("hat(mu) == ", base, " %*% 10^", expo)
}

p1 <- ggplot(site_comparison_cat, aes(x = Project_SampleType, y = Spores.m3)) +
  geom_point(data = fasmee23_k4_spores,
             aes(x = Project_SampleType, y = TotalSpores_corrected.m3),
             position = position_jitter(width = 0.1),
             alpha = 0.5, size = 2, color = "#0072B2") +
  geom_violin(data = fasmee23_k4_spores,
              aes(x = Project_SampleType, y = TotalSpores_corrected.m3),
              fill = NA, color = "black", width = 0.8, alpha = 0.5) +
  geom_col(fill = "#0072B2", alpha = 0.6) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                width = 0.2, color = "black", linewidth = 1.1) +
  labs(x = "Site and Sample Type",
       y = expression(Spores~m^{-3}),
       title = "Fungal Spore Abundance: Site Comparison") +
  theme_bw() +
  theme(text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = x, xend = x, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_segment(data = sig_bars,
               aes(x = xend, xend = xend, y = y, yend = y - max_height * 0.05),
               linewidth = 0.5) +
  geom_text(data = sig_bars,
            aes(x = (x + xend) / 2, y = y + max_height * 0.02, label = annotation),
            size = 3.2, vjust = 0, parse = TRUE) +
  scale_x_discrete(labels = c("Konza4\nAmbient", "Konza4\nSmoke",
                              "FASMEE23\nAmbient", "FASMEE23\nSmoke")) +
  geom_label(
    data = site_comparison_cat,
    aes(
      x = as.numeric(Project_SampleType) + 0.16,
      y = Spores.m3,
      label = sapply(Spores.m3, sci_label)
    ),
    parse = TRUE,
    hjust = 0,
    vjust = -0.2,
    fill = "white",
    color = "black",
    label.size = 0.3,
    fontface = "plain",
    size = 2.5
  ) +
  coord_cartesian(ylim = c(0, max(fasmee23_k4_spores$TotalSpores_corrected.m3 * 0.5)))

print(p1)
```

## [14.4] Comparison of Konza4 and FASMEE23 Results

### Statistical Results Comparison

| Site     | Low Smoke      | High Smoke     |
|----------|----------------|---------------|
| Konza4   | 5.35× (p=0.037)| 5.28× (p=0.030)|
| FASMEE23 | 6.06× (p<0.0001)| 3.02× (p=0.016)|

| Site     | None   | Low     | High    |
|----------|--------|---------|---------|
| Konza4   | 19,780 | 105,774 | 104,491 |
| FASMEE23 | 81,232 | 492,328 | 245,717 |

### Key Findings
- **Baseline Differences:** FASMEE23 has ~4× higher ambient spore concentrations than Konza4, indicating fundamental ecosystem differences.
- **Smoke Response Patterns:** Konza4 shows similar increases for both low and high smoke (~5.3×), while FASMEE23 shows a stronger response to low smoke (6.06×) than high smoke (3.02×).
- **Absolute Concentrations:** FASMEE23 reaches much higher peak concentrations in smoke (up to ~492,000 spores/m³).
- **Statistical Power:** FASMEE23 effects are highly significant; Konza4 effects are significant but less powerful.

### Model Quality Assessment
- Both models show excellent diagnostics, robust random effects, and strong evidence for smoke effects.
- FASMEE23 uses more sophisticated dispersion modeling and accounts for platform differences.

### Ecological Interpretations
- Forest systems (FASMEE23) have higher baseline fungal abundance and more variable smoke response.
- Prairie systems (Konza4) show a more consistent dose-response to smoke.
- Combustion and ecosystem differences likely drive these patterns.

### Limitations
- Point-in-time sampling; no seasonal/long-term trends captured.
- Localized drone sampling may not represent landscape-scale patterns.
- Some model overparameterization and unbalanced designs.
- Measurement uncertainty in spore viability and PM2.5-based smoke classification.

### Conclusions
Both GLMMs provide robust evidence for smoke effects on fungal spore concentrations, with forest fires producing higher absolute concentrations and prairie fires showing more consistent dose-response. The modeling approach effectively captures bioaerosol dynamics across fire regimes, though temporal and spatial limitations should be considered.




